The white-footed deermouse, an infection tolerant reservoir for several zoonotic agents, tempers interferon responses to endotoxin in comparison to the mouse and rat

Version 1

Abstract

Abstract

The white-footed deermouse Peromyscus leucopus, a long-lived rodent, is a key reservoir for agents of several zoonoses, including Lyme disease. While persistently infected, this deermouse is without apparent disability or diminished fitness. For a model for inflammation elicited by various pathogens, the endotoxin lipopolysaccharide (LPS) was used to compare genome-wide transcription in blood by P. leucopus, Mus musculus and Rattus norvegicus and adjusted for white cell concentrations. Deermice were distinguished from the mice and rats by LPS response profiles consistent with non-classical monocytes and alternatively-activated macrophages. LPS-treated P. leucopus, in contrast to mice and rats, also displayed little transcription of interferon-gamma and lower magnitude fold-changes in type 1 interferon-stimulated genes. This was associated with comparatively reduced transcription of endogenous retrovirus sequences and cytoplasmic pattern recognition receptors in the deermice. The results reveal a mechanism for infection tolerance in this species and perhaps other animal reservoirs for agents of human disease.

Introduction

How does the white-footed deermouse Peromyscus leucopus continue to thrive while sustaining infections with disease agents it serves as reservoir for (Barbour 2017)? The diverse tickborne pathogens (and diseases) for humans include the extracellular bacterium Borreliella burgdorferi (Lyme disease), the intracellular bacterium Anaplasma phagocytophilum (anaplasmosis), the protozoan Babesia microti (babesiosis), and the Powassan flavivirus (viral encephalitis). Most deermice remain persistently infected but display scant inflammation in affected tissues (Moody et al. 1994; Cook and Barbour 2015; Long et al. 2019), and without apparent consequence for fitness (Schwanz et al. 2011; Voordouw, Lachish, and Dolan 2015).

A related question—conceivably with the same answer--is what accounts for the two-to-three fold longer life span for P. leucopus than for the house mouse, Mus musculus (Sacher and Hart 1978; Labinskyy et al. 2009)? The abundance of P. leucopus across much of North America (Hall 1979; Moscarella et al. 2019) and its adaptation to a variety of environments, including urban areas and toxic waste sites (Biser et al. 2004; Levengood and Heske 2008; Munshi-South and Kharchenko 2010), indicates successful adjustment to changing landscapes and climate. Peromyscus species, including the hantavirus reservoir P. maniculatus (Morzunov et al. 1998), are more closely related to hamsters and voles in family Cricetidae than to mice and rats of family Muridae (Bradley et al. 2014).

As a species native to North America, P. leucopus is an advantageous alternative to the house mouse, which is of Eurasian origin, for study of natural variation in populations that are readily at-hand (Bedford and Hoekstra 2015; Long et al. 2022). A disadvantage is the limited reagents and genetic tools of the sorts that are applied for mouse studies. As an alternative, we study P. leucopus with a non-reductionist approach that is comparative in design and agnostic in assumptions (Balderrama-Gutierrez et al. 2021). The genome-wide expression comparison for P. leucopus is with M. musculus and, added here, the brown rat Rattus norvegicus. Given the wide range of pathogenic microbes that deermice tolerate, we use the bacterial endotoxin lipopolysaccharide (LPS) as the primary experimental treatment because the inflammation it elicits within a few hours has features common to different kinds of serious infections, not to mention severe burns and critical injuries (Xiao et al. 2011).

We previously reported that a few hours after injection of LPS, P. leucopus and M. musculus had distinguishing profiles of differentially expressed genes (DEG) in the blood, spleen, and liver (Balderrama-Gutierrez et al. 2021). In brief, the inflammation phenotype of deermice was consistent with an “alternatively-activated” or M2-type macrophage polarization phenotype instead of the expected “classically-activated” or M1-type polarization phenotype that was observed for M. musculus (Murray et al. 2014). The deermice also differed from mice in displaying evidence of greater neutrophil activation and degranulation after LPS exposure. The potentially damaging action from neutrophil proteases and reactive oxygen species appeared to be mitigated in part in P. leucopus by proteins like secretory leukocyte protease inhibitor (Slpi) and superoxide dismutase 2 (Sod2).

Here, we first address whether the heightened transcription of neutrophil-associated genes in P. leucopus is attributable to differences in numbers of white cells in the blood. To better match for genetic diversity, we substituted outbred M. musculus for the inbred BALB/c mouse of the previous study. We retained the experimental protocol of short-term responses to LPS. This main experiment was supplemented by a study of rats under the similar conditions and by an investigation of a different dose of LPS and duration of exposure in another group of deermice. The focus was on the blood of these animals, not only because the distinctions between species in their transcriptional profiles were nearly as numerous for this specimen as for spleen and liver (Balderrama-Gutierrez et al. 2021), but also because for ecological and immunological studies of natural populations of Peromyscus species blood is obtainable from captured-released animals without their sacrifice.

The results inform future studies of Peromyscus species, not only with respect to microbial infections and innate immunity, but also determinants of longevity and resilience in the face of other stressors, such as toxic substances in the environment. The findings pertain as well to the phenomenon of infection tolerance broadly documented in other reservoirs for human disease agents, such as betacoronaviruses and bats (Mandl et al. 2018). Less directly, the results provide for insights about maladaptive responses among humans to microbes, from systemic inflammatory response syndrome (SIRS) to post-infection fatigue syndromes.

Results

LPS experiment and hematology studies

Twenty adult animals of each species and equally divided between sexes received by intraperitoneal injection either purified E. coli LPS at a dose of 10 µg per g body mass or saline alone (Table 1). Within 2 h LPS-treated animals of both species displayed piloerection and sickness behavior, i.e. reduced activity, hunched posture, and huddling. By the experiment’s termination at 4 h, 8 of 10 M. musculus treated with LPS had tachypnea, while only one of ten LPS-treated P. leucopus displayed this sign of the sepsis state (p = 0.005).

Within a given species there was little difference between LPS-treated and control animals in values for erythrocytes. But overall the deermice had lower mean (95% confidence interval) hematocrit at 42% (36-48), hemoglobin concentration at 13.8 g/dL (12.1-15.5), and mean corpuscular volume for erythrocytes at 49 fL (47-51) than M. musculus with respective values of 56% (51-62), 16.1 g/dL (14.6-17.7), and 60 fL (58-62) (p < 0.01). These hematology values for adult CD-1 M. musculus and LL stock P. leucopus in this study were close to what had been reported for these colony populations (CharlesRiver 2012; Wiedmeyer et al. 2014).

In contrast to red blood cells, the mean numbers of white blood cells in the LPS groups in both species were lower than those of control groups (Figure 1). Controls had a mean 4.9 (3.5-6.4) x 103 white cells per µl among M. musculus and 5.8 (4.2-7.4) x 103 white cells per µl among P. leucopus (p = 0.41). For the LPS-treated animals the values were 2.1 (1.5-2.7) x 103 for mice and 3.1 (0.9-5.4) x 103 for deermice (p = 0.39). However, there was difference between species among LPS-treated animals in the proportions of neutrophils and lymphocytes in the white cell population. The ratios of neutrophils to lymphocytes were 0.25 (0.14-0.45) and 0.20 (0.13-0.31) for control M. musculus and P. leucopus, respectively (p = 0.53). But under the LPS condition. the neutrophil-to-lymphocyte ratio was 0.18 (0.11-0.28) for mice and 0.64 (0.42-0.97) for deermice (p = 0.0006). The regression curves for plots of neutrophils and lymphocytes for LPS-treated and control P. leucopus and LPS-treated M. musculus had similar slopes, but the y-intercept was shifted upwards towards a higher ratio of neutrophils to lymphocytes for blood from the LPS group of deermice. Control group mice and deermice and LPS-treated mice had similar percentages (∼5%) of monocytes in their blood; the mean monocyte percentage rose to 10% in LPS treated deermice (p = 0.12). Eosinophil percentages tended to be higher in deermice at a mean 3.4 (2.1-4.7)% than mice at 1.2 (0.5-1.9)% under either condition (p = 0.004).

Total white blood cells, neutrophils, and lymphocytes of Mus musculus (M) and Peromyscus leucopus (P) with or without (control; C) treatment with 10 µg lipopolysaccharide (LPS; L) per g body mass 4 h previous. The box plots of left and center panels show values of individual animals and compile median, quartiles, and extreme values. The linear regressions of the right panel are color-coded according to the species and treatment designations. The outlier value for a M. musculus control was excluded from the linear regression for that group.

In the P. leucopus experiment with a 10-fold lower dose of LPS and a 12 hour duration, the mean (95% confidence interval) white blood cell count (x 103) at termination 3.5 (2.5-4.5) in controls and 7.9 (6.0-9.7) in the LPS-treated (p = 0.01). Even with the higher overall white blood cell count the increase white cells was proportionately higher for neutrophils than for lymphocytes, as was seen in the deermice in the higher dose LPS experiment. The ratio of neutrophils-to-lymphocytes was 0.20 (0.07-0.32) in the controls and 0.38 (0.26-0.50) (p = 0.10).

The higher neutrophil to lymphocyte ratio in the deermice exposed to LPS was consistent with the greater neutrophil activation noted by transcriptional analysis (Balderrama-Gutierrez et al. 2021). But many individual genes that constitute this and related gene ontology (GO) terms had transcription levels in the deermice that far exceeded a three-fold difference in neutrophil counts. For some genes fold-changes were a hundred or more fold, which suggested that the distinctive LPS transcriptional response profile between species was not attributable solely to neutrophil counts.

Genome-wide expression in blood of deermice and mice

We used the respective transcript sets from the reference genomes for P. leucopus and M. musculus for deep coverage RNA-seq with paired-end ∼150 nt reads. Principle component analyses (PCA) of the P. leucopus data and M. musculus data revealed that untreated controls had coherent profiles within each species (Figure 2). With the exception of one mouse, the LPS-treated M. musculus were also in a tight PCA cluster. In contrast, the LPS-treated deermice displayed a diversity of genome-wide transcription profiles and limited clustering.

Principle component analysis of genome-wide RNA-seq data of P. leucopus or M. musculus with (red dot) or without (blue dot) treatment with LPS 4 h previous. The individual animals listed in Table 1 are indicated on the graphs.

For both species the number of genes with higher expression with LPS exposure exceeded those with lower or unchanged expression (Dryad Table D1). For P. leucopus and M. musculus the mean fold-changes were 1.32 (1.29-1.35) and 1.30 (1.24-1.36), respectively (p = 0.31). Among P. leucopus 1154 DEGs with an absolute fold-change ≥ 2 and false-discovery rate p value < 0.003 were identified. The corresponding number for M. musculus was 1266. Two DEGs that distinguished females from males among controls or LPS-treated were the non-coding Xist, the X chromosome inactivating gene, for females and the Y chromosome-linked DEAD-box helicase 3 (Ddx3y) gene for males.

Figure 3 shows the GO term analyses of these DEGs for P. leucopus and M. musculus. The up-regulated gene profile for P. leucopus featured GO terms associated with “neutrophil degranulation”, “myeloid leukocyte activation”, “leukocyte migration”, and “response to molecule of bacterial origin”. Other sets of up-regulated genes for the deermice were “negative regulation of cytokine production” and “regulation of reactive oxygen species metabolic process”. None of these were up-regulated gene GO terms with p values < 1011 for M. musculus. Indeed, “leukocyte activation” and “leukocyte migration” were GO terms for down-regulated DEGs in M. musculus. Distinctive GO terms for up-regulated genes distinguishing mice from deermice were “response to virus”, “response to interferon-beta”, “response to interferon-gamma”, “response to protozoan”, and “type II interferon signaling”.

Gene Ontology (GO) terms associated with up-regulated genes (upper panels) and down-regulated genes (lower panels) of P. leucopus and M. musculus treated with LPS in comparison with untreated controls of each species.

Of the top 100 DEGs for each species by ascending FDR p value, 24 genes were shared between species (Supplementary Materials Table S1). These included up-regulated Bcl3, Ccl3, Cxcl1, Cxcl2, Cxcl3, Cxcl10, Il1rn, and Sod2. Among the 100 mouse DEGs, 20 were constituents of GO terms “response to virus” (GO:0009615) or “response to interferon-beta” (GO:0035456) and only 3 were members of GO term sets “response to molecule of bacterial origin” (GO:0002237) or “response to lipopolysaccharide” (GO:0032496). In contrast, among the top 100 deermouse DEGs, there were only 2 associated with the virus or type 1 interferon GO terms, but 12 were associated with either or both of the bacterial molecule GO terms.

The emerging picture was of P. leucopus responding to LPS exposure as if infected with an extracellular bacterial pathogen, including with activated neutrophils. While M. musculus animals shared with P. leucopus some features of an antibacterial response, they also displayed type I and type II interferon type response profiles associated with infections with viruses and intracellular bacteria and parasites.

Targeted RNA seq analysis

A challenge for cross-species RNA-seq is commensurability between annotated transcripts of reference sets. Orthologous genes can be identified, but mRNA isoforms and their 5’ and 3’ untranslated regions may not fully correspond. Accordingly, we limited targeted RNA-seq to protein coding sequences of mRNAs for the corresponding sets of P. leucopus and M. musculus sequences.

For cross-species normalization we chose a gene whose transcription approximated the varying numbers of white cells. The myeloid cell marker CD45, formally “protein tyrosine phosphatase, receptor type, C”, and encoded by the Ptprc gene, was the denominator transcript for normalization across species and treatments. In humans, mice, and hamsters, Ptprc is expressed by nucleated hematopoietic cells and used as a white cell marker for flow cytometry (Schnizlein-Bick et al. 2002). Pearson’s continuous and Spearman’s ranked correlations between log-transformed total white blood cell counts and normalized reads for Ptprc across 40 animals representing both species, sexes, and treatments were 0.40 (p = 0.01) and 0.34 (p = 0.03), respectively.

The 113 mRNA coding sequences, which are listed in Methods, were drawn from the identified DEGs for P. leucopus and M. musculus from the genome-wide RNA-seq. Figure 4 comprises plots of the log-transformed mean target-to-Ptprc ratios for the 10 P. leucopus controls and 10 M. musculus controls (left panel) and for the 10 P. leucopus and 10 M. musculus treated with LPS (right panel). For untreated animals there was high correlation and a regression coefficient of ∼1 between the paired data for deermice and mice. The mitochondrial cytochrome oxidase 1 gene (MT-Co1) and S100a9, a subunit of calprotectin, were comparably transcribed. But, there were other coding sequences that stood out for either their greater or lesser transcription in untreated deermice than mice. Two examples of greater expression were Arg1 and MX dynaminin-like GTPase 2 (Mx2), an ISG, while two examples of lesser expression were matrix metalloprotease 8 (Mmp8) and Slpi. There was low to undetectable transcription of Nos2 and interferon-gamma (Ifng) in the blood of controls of both species.

Scatter plots with linear regression of pairs of log-transformed (ln) normalized RNA-seq reads for selected coding sequences for control P. leucopus and M. musculus (left panel) and LPS-treated P. leucopus and M. musculus (right panel). The coefficients of determination (R2) and selected genes (each by a different symbol) are indicated in each graph.

For the LPS-treated animals there was, as expected for this selected set, higher expression of the majority genes and greater heterogeneity between P. leucopus and M. musculus in their responses for represented genes. In contrast to the findings with controls, Ifng and Nos2 were substantially more transcribed in treated mice, but these genes were only marginally so in deermice. A comparatively restrained transcriptional response in deermice was also noted for Mx2. On the other hand, there were greater fold-change from baseline in P. leucopus than in M. musculus for interleukin-1 beta (Il1b), Mmp8, Slpi, and S100a9.

Table 2 lists all the selected targets with the means and confidence intervals for the normalized values for controls and LPS-treated M. musculus and controls and LPS-treated P. leucopus, as well as the fold-changes within each species and between treatments.The values for individual animals by species, sex, and treatment are provided in Supplementary Materials Table S2. In Table 2. The final column is the ratio of the fold-change between LPS to control in P. leucopus to the corresponding value for M. musculus. This along with the derived heat-map of these ratios, which presented in the second column, indicates the genes for which there was little difference between species in their responses to LPS--either up or down--as well as those that were comparatively greater or lesser in one species or the other. Several of these genes are considered in other specific contexts below.

To assess differences by sex among P. leucopus in their responses to LPS, we used the original reads from 20 deermice from the previous blood RNA-seq study in combination with the data for 20 deermice of the present study for the targeted RNA-seq. This yielded 11 deermice of each sex receiving LPS in the same dose and 9 animals of each sex receiving saline instead. Using Z-scores to provide comparability across studies and after correcting for multiple testing, we found no difference between females and males in their responses, beyond a tendency for Il6 and Il10 expression to be higher in males than females in the LPS treatment group.

“Alternatively-activated” macrophages and “nonclassical” monocytes in P. leucopus

While we could not type single cells using protein markers, we could assess relative transcription of established indicators of different white cell subpopulations in whole blood. The present study, which incorporated outbred M. musculus instead of an inbred strain, confirmed the previous finding of an inversion between Nos2 and Arg1 expression: that is, high Nos2 and low Arg1 in M. musculus and the opposite in P. leucopus (Figure 4; Table 2). The RNA-seq findings were confirmed by specific RT-qPCR assays for Nos2 and Arg1 transcripts for P. musculus and M. musculus (Table 3).

Low transcription of Nos2 and high transcription of Arg1 both in controls and LPS-treated P. leucopus was also observed in the experiment where the dose of LPS was 1 µg/g body mass instead of 10 µg/g and the interval between injection and assessment was 12 h instead of 4 h (Table 4). We previously showed that the Nos2-Arg1 relationship in deermice was not limited to response to LPS; P. leucopus that were bacteremic with the relapsing fever agent Borrelia hermsii had no detectable transcripts for Nos2 but a higher amount of transcription of Arg1 in the blood in comparison to uninfected animals (Balderrama-Gutierrez et al. 2021).

In addition to the ratio of Nos2 to Arg1 expression for typing macrophage and monocyte subpopulations, there are also the relative expressions of three other pairs of genes: (1) IL12 and IL10, where a lower IL12/IL10 ratio is more characteristic of alternatively activated or M2 type (van Stijn et al. 2015; P.J. Murray 2017); (2) the proto-oncogene kinases Akt1 and Akt2, where the associations are Akt1 with M2-type and Akt2 with M1-type macrophages (Vergadi et al. 2017; Arranz et al. 2012); and (3) CD14 and CD16 (low affinity immunoglobulin gamma Fc region receptor III; Fcgr3), where low expression of CD14 and high expression of CD16/Fcgr3 is associated with “non-classical” monocytes (Narasimhan et al. 2019). There is evidence that nonclassical monocytes can change to M2-type macrophages (Italiani and Boraschi 2014).

These four relationships, which are presented as log-transformed ratios of Nos2/Arg1, IL12/IL10, Akt1/Akt2, and CD14/CD16, are shown in Figure 5. The differences between P. musculus and M. musculus in the ratios of Nos2/Arg1 and IL12/IL10 were reported before (Balderrama-Gutierrez et al. 2021), and the present study with outbred mice and normalization for white cells replicated those findings. In both species the Akt1/Akt2 ratio declined in LPS-treated animals, but for P. leucopus the ratio remained > 1.0 even among LPS-treated animals, while in the blood of M. musculus the ratio was < 1.0 at baseline and declined further in the LPS-treated animals.

Box plots of log-transformed ratios of four pairs of gene transcripts from targeted RNA-seq analysis of blood of P. leucopus (P) or M. musculus (M) with (L) or without (C) treatment with LPS.

An ortholog of Ly6c (Bothwell, Pace, and LeClair 1988), a protein used for typing mouse monocytes and other white cells, has not been identified in Peromyscus or other Cricetidae family members, so the comparison with Cd14 is with Cd16 or Fcgr3, which cricetines like humans do have. In mice the Cd14/Cd16 ratio increased from baseline in the LPS group. In the deermice the ratio in control animals was midway between the two groups of mice but there was a marked decrease in the LPS-treated deermice (Figure 5). This was not associated with a fall in the absolute numbers or percentages of monocytes in the blood of these animals.

Taken together, the Nos2-Arg1, Il12-Il10, Akt1-Akt2, and CD14-CD16 relationships document a disposition toward alternatively-activated macrophages and nonclassical monocytes in P. leucopus both before and after exposure to LPS. This contrasts with profiles consistent with a predominance of classically-activated macrophages and classical monocytes in mice.

Interferon-gamma and interleukin-1 beta dichotomy between deermice and murids

For mice the Ifng transcript was one of the top ranked DEGs by both fold-change and adjusted p value by genome-wide RNA-seq (Table 2). In contrast, for P. leucopus Ifng was far down the list and the comparably ranked DEG instead was Il1b. This inversion of relationships between two pro-inflammatory cytokines was confirmed by analysis of the individual animals of both species (Figures 6). There was little or no detectable transcription of Ifng in the blood of deermice in which Il1b expression was high. There was also scant to no transcription of Ifng in the blood of P. leucopus 12 h after injection of LPS (Table 4).

Transcripts of interferon-gamma (γ) and interleukin-1 beta (β) by targeted RNA-seq of the blood of P. leucopus (P) or M. musculus (M) with (L) or without (C) treatment with LPS. The top panels are box plots of the individual values. The lower left panel is a scatter plot of interleukin-1 β on interferon-γ values. The lower right panel is a Discriminant Analysis of these pairs of values where Factor 1 corresponds to interferon-γ, and Factor 2 corresponds to interleukin-1 β.

The up-regulation of Ifng within 4 hours of exposure to LPS was not limited to the species M. musculus. In an experiment with the rat R. norvegicus, we used two different LPS doses (5 µg/g and 20 µg/g) but the same 4 h endpoint and whole blood as the sample. Both groups of LPS-treated rats had lowered total white blood cells and, like the mice, lower neutrophil-to-lymphocyte ratios compared to controls (Table 5). There were also marked elevations of interferon-gamma, interleukin-6 and interleukin-10 proteins from undetectable levels in the blood of the treated rats. By targeted RNA-seq there were 24x fold-changes between the LPS-treated rats and control rats for Ifng and Nos2 but only ∼3x fold-change for Il1b. We also noted that in the rat the Nos2-Arg1, Il12-Il10, Akt1-Akt2, and CD14-CD16 relationships that serve to signify monocyte and macrophage phenotypes resembled those of the mice and not what was observed in deermice.

We asked why the interferon-gamma response observed in CD-1 mice and rats here was not as pronounced in BALB/c mice (Balderrama-Gutierrez et al. 2021). Accordingly, we used the RNA-seq reads from the prior study in combination with the reads of the present study and carried out targeted RNA-seq as described (Supplementary Materials Figure S1). The BALB/c inbred mice had, like the CD-1 mice, modest elevations of Il1b transcription. Ifng expression was also elevated in the BALB/c animals but not to the degree noted in CD-1 mice or rats. An explanation for this may be an inherent difference of BALB/c mice from other strains in their lower interferon-gamma response to LPS (Soudi et al. 2013; Kuroda, Kito, and Yamashita 2002).

Interferon-gamma and inducible nitric oxide synthase

Interferon-gamma is a determinant of Nos2 expression (Lowenstein et al. 1993; Salkowski et al. 1997). So, the scant transcription of Ifng in P. leucopus conceivably accounted for the low expression of Nos2 in that species, not only with LPS but also during bacterial infection. The analysis shown in upper left panel of Figure 7 shows a tight correlation between the levels of transcription of Ifng and Nos2 for both species and both experimental conditions. A significant correlation was also observed for the combined set of animals between the ratios of Nos2 to Arg1 and Ifng to Il1b (upper right panel), an indication of co-variation between Ifng expression and macrophage polarization.

Normalized transcripts of Nos2, Ifng, and Cd69 in targeted RNA-seq analysis of blood of P. leucopus (P) or M. musculus (M) with (L) or without (C) treatment with LPS. Upper left: scatter plot of individual values for Nos2 on Ifng with linear regression curve and coefficient of determination (R2). Upper right: natural logarithm (ln) of ratios of Nos2/Argi1 on Ifng/IL1b with regression curve and R2. Lower left: Box plots of individual values of normalized transcripts of Cd69. Lower right: Scatter plot of Ifng on Cd69 with separate regression curves and R2 values for M. musculus and P. leucopus.

The plausible sources of Ifng mRNA in whole blood are T-cells, Natural Killer (NK) cells, and Type 1 Innate Lymphoid Cells (ILC1) (Quatrini et al. 2017). A DEG for M. musculus by both genome-wide and targeted RNA-seq (Table 2; Table S2) was Cd69, a C-type lectin protein and an early activation antigen for these cells (Heinzelmann et al. 2000). In P. leucopus transcription Cd69 occurred in the blood of control P. leucopus but it was the same or only marginally different for the LPS-treated animals (lower left of Figure 7). In contrast, in M. musculus the baseline transcription of Cd69 was below that of P. leucopus, but in the LPS-treated mice it was many fold higher. In mice transcripts for Cd69 correlated tightly with Ifng transcription, while in the deermice there was little correlation between Cd69 and Ifng expression at those low levels (lower right).

The findings are consistent with CD69-positive cells being a source of Ifng in mice. Cd69 transcription was comparatively higher in control deermice than in control mice, so we presume that deermice have CD69-positive cells at baseline. One explanation then for the comparatively few Ifng transcripts in the deermice after LPS is a diminished responsiveness of these cells. Tlr4 expression increased ∼3-fold more in P. leucopus than in M. musculus after LPS, but the magnitude of the relative decline in expression of Cd14 in deermice than mice was even greater (Table 2). CD14 is required for the LPS-stimulated signaling through surface TLR4 (Mazgaeen and Gurung 2020), and, as such, its decreased availability for this signaling pathway is a possible explanation for the moderated response to LPS in P. leucopus.

Interferon-stimulated genes and RIG-I-like receptors

As noted, GO terms differentiating mice from deermice included “response to interferon-beta” and “response to virus” (Figure 3). There was also the example of Mx2, an ISG with anti-viral activity on its own, that showed a greater fold-change from baseline in mice than in deermice (Figure 4 and Table 2). Five other ISGs--guanylate binding protein 4 (Gbp4), interferon-induced protein with tetratricopeptide repeat (Ifit1), interferon regulatory factor 7 (Irf7), ubiquitin-type modifier ISG15 (Isg15), and 2’-5’ oligoadenylate synthase 1A (Oas1a)—had elevated transcription in all LPS-treated animals. But the magnitude of fold change was less in the deermice, ranging from 6-25% of what it was in the LPS group of mice (Table 2).

The up-regulation of these ISGs was evidence of type 1 interferon action, but transcripts for interferon-1 beta (Ifnb) or -alpha (Ifna) themselves were scarcely detectable in deermice or mice in the blood under either condition. We then considered pattern recognition receptors (PRR) that conceivably part of a signaling pathway leading to ISG expression. Among the DEGs from the genome-wide analyses were four cytoplasmic PRRs (Table S2): (1) Rigi (formerly Ddx58), which encodes the RNA helicase retinoic acid-inducible I (RIG-I); (2) Ifih1, which encodes interferon induced with helicase C domain 1, also known as MDA5 and a RIG-I-like receptor; (3) Dhx58, also known LGP2 and another RIG-I-like receptor; and (4) cGAS (cyclic GMP-AMP synthase), which is part of the cGAS-STING sensing pathway.

All four of these cytoplasmic PRRs were up-regulated in blood of LPS-treated mice and deermice (Table 2). But, again, for each of them the magnitude of fold change was less by 50-90% in treated P. leucopus than in M. musculus. The coefficients of determination for the 6 ISGs and the 4 PRRs are provided in Figure 8. For most of the pairs there was evidence of covariation across all 40 animals. When the correlation was low across all the data, e.g. between the ISG Mx2 and the PRR Rigi or the ISGs Mx2 and Gbp4, it was high within a species.

Co-variation between transcripts for selected PRRs (yellow) and ISGs (gree) in the blood of P. leucopus (P) or (M) with (L) or without (C). Top panel: matrix of coefficients of determination (R2). Bottom panels: scatter plots of log-transformed normalized Mx2 transcripts on Rigi (left), Ifih1 (center), and Gbp4 (right). The linear regression curves are for each species. For the right bottom graph the results from the General Linear Model (GLM) estimate are given.

These findings were evidence that pathways in P. leucopus from PRR signaling and ISG expression functioned similarly to those in M. musculus but differed under these experimental conditions in magnitude of the changes, being more moderate in the deermice.

Endogenous retroviruses in deermice, mice, and rats after LPS exposure

The six ISGs are nonexclusive consequences of activity of type I interferons. What we could document was the association of transcription of the gene for the cytoplasmic PPRs, including RIG-I, and the ISGs in both species, as well as the distinction between deermice in the magnitude of the responses of both PRRs and ISGs. These findings led us to ask what could be a pathogen-associated molecular pattern (PAMP) for signaling pathways leading to expression of type 1 interferons.

One of these is endogenous retroviruses (ERV). The activity of these diverse, abundant, and pervasive elements have been recognized as one of the drivers of innate immune responses to a microbe (Hurst and Magiorkinis 2015; Lima-Junior et al. 2021; Rangel et al. 2022). Our attention was drawn to ERVs by finding in the genome-wide RNA-seq of LPS-treated and control rats that 2 of the 3 highest scoring DEGs by FDR p value and fold-change were a gag-pol polyprotein of a leukemia virus with 131x fold-change from controls and a mouse leukemia virus (MLV) envelope (Env) protein with 62x fold-change (Table 5).

We returned to the mouse and deermouse data. There were four MLV or other ERV Env proteins among the 1266 genome-wide RNA-seq DEGs for M. musculus. But, there was no ERV Env protein identified as such among the 1154 DEGs identified for P. leucopus (Table S1; Dryad Table D1). One possible explanation for the difference was an incomplete annotation of the P. leucopus genome. We took three approaches to rectify this. The first was to examine the DEGs for P. leucopus that encoded a polypeptide ≥ 200 amino acids and was annotated for the genome as “uncharacterized”. A search with these candidates of both the virus and rodent proteins databases identified two that were homologous with gag-pol polyproteins of ERVs, mainly leukemia viruses, of mammals.

For a second approach we carried out a de novo transcript assembly of mRNA reads from blood of LPS-treated and control P. leucopus and used the resultant contigs as the reference set for RNA-seq analysis. This identified two contigs that were measurably transcribed in the blood, differentially expressed between conditions, and were homologous to ERV sequences. One was revealed to be an Env protein that was identical to a P. leucopus protein annotated as “MLV-related proviral Env protein” (XP_037065362). The second was a gag-pol protein. The latter was near-identical to the gag-pol protein identified by the first approach.

The third approach was to scan the P. leucopus genome for nonredundant sequences, defined as < 95% identity, that were homologous with ERV pol sequences, which are not typically annotated because of masking for repetitive sequences This analysis yielded 615 unique sequences. These were used in turn as reference set for RNA-seq. There were 4 sequences that met the criterion of FDR p value <0.05. Three were transcribed at 5- to 40-fold higher levels in LPS-treated deermice than in controls. But all three, as well as the fourth, a down-regulated DEG, were ERV relics with truncations, frame shifts, and in-frame stop codons. These were assessed as non-coding RNAs and not further pursued in this study.

To represent P. leucopus in a targeted RNA-seq comparison with mice and rats we settled on the Env protein and gag-pol coding sequences identified present in the blood mRNA and as DEGs. Representing M. musculus were highest ranked MLV Env and gag-pol protein sequences among the DEGs. For rats we chose Env and gag-pol proteins that were second and third ranked DEGs identified in the genome-wide RNA-seq (Dryad table). Because of length differences for the coding sequences, the unit used for cross-species analysis was reads per kilobase before normalization for Ptprc transcription.

The left panel of Figure 9 shows the striking transcriptional fold-change in LPS-treated rats of both Env and gag-pol sequences over controls. Of lesser magnitude but no less significant was the fold-change observed M. musculus for both Env and gag-pol sequences. In both mice and rats Env and gag-pol read values were highly correlated across conditions. In contrast, in P. leucopus the magnitudes of fold-change upwards for gag-pol was less than in mice or rats, and transcription of the Env protein sequence was actually lower in LPS-treated animals than in controls. While there was a tight association between Env protein and the PRR Rigi transcription in the M. musculus, this was not observed in P. leucopus. Rigi transcription was moderately higher at the time that Env protein’s transcription was lower in the LPS group (right panel of Figure 9).

Scatter plots of endogenous retrovirus (ERV) Env and Gag-pol protein gene transcription (left) and association of ERV Env with Rigi transcription (right) in the blood of P. leucopus (Pero; P), M. musculus (Mus; M), or R. norvegicus (Rattus) with (L) or without (control; C) treatment with LPS. In right panel the linear regression curve and coefficients of determination (R2) for P. leucopus and M. musculus are shown.

Discussion

Study limitations

The approach was forward and unbiased, looking for differences between species broadly across their transcriptomes. The findings lead to hypotheses, but reverse genetics in service of that testing was not applied here. In selective cases we could point to supporting evidence in the literature on M. musculus and the phenotypes of relevant gene knockouts, but there are no such resources for Peromyscus as yet. The resource constraint also applies to the availability of antibodies for use with Peromyscus for immunoassays for specific proteins, e.g. interferon-gamma, in serum, or for cell markers, e.g. CD69, for flow cytometry of white blood cells.

While a strength of the study was use of an outbred population of M. musculus to approximate the genetic diversity of the P. leucopus in the study, this meant that some genes of potential relevance might have gone undetected, i.e. from type II error. The variances for a sample of genetically diverse outbred animals, like the LL stock of P. leucopus (A. Long et al. 2019; P. Long et al. 2022), would be expected to be greater than for the same sized sample of inbred animals. For some traits, especially ones that are complex or under balancing selection, even sample sizes of 10 in each group may not have provided sufficient power for discrimination between deermice and mice.

The parameters for the experiment of LPS dose, the route, and duration of experiment each might have had different values under another design. Those particular choices were based on past studies of deermice and mice (Langeroudi et al. 2014; Balderrama-Gutierrez et al. 2021). In another experiment we found that with doses twice or half those given the deermice the responses by rats to the different doses were indistinguishable by hematology, cytokine assays, and RNA-seq. Thus, there seems to be some latitude in the dose and still achieving replication. We obtained similar results for P. leucopus when we looked at a replicate of the experiment with the same conditions (Balderrama-Gutierrez et al. 2021), when the dose was lower and duration lengthened to 12 h (this study), and when the inflammatory stimulus was bacteremia (Barbour et al. 2019). The bacterial infection experiment indicated that the observed effect in P. leucopus was not limited to a TLR4 agonist; the lipoproteins of B. hermsii are agonists for TLR2 (Salazar et al. 2009).

While the rodents in these experiments were housed in the same facility and ate the same diet, we cannot exclude inherent differences in gastrointestinal microbiota between species and individual outbred animals as co-variables for the experimental outcomes. We reported differences between the LL stock P. leucopus and BALB/c M. musculus of the same age and diet in their microbiomes by metagenomic analysis and microbiologic means (Milovic et al. 2020). This included a commensal Tritrichomonas sp. in P. leucopus but not in the M. musculus in the study. The presence of these protozoa affects innate and adaptive immune responses in the gastrointestinal tract (Escalante et al. 2016; Chiaranunt et al. 2022), but it is not clear whether there are systemic consequences of colonization by this flagellate.

LPS, ERVs, and interferons

The results confirm previous reports of heightened transcription of ERV sequences in mice or mouse cells after exposure to LPS (Hara et al. 1981; Jongstra and Moroni 1981; Stoye and Moroni 1983). Here we add the example of the rat. The LPS was administered in solution and not by means of membrane vesicles. The sensing PRR presumably was surface-displayed, membrane-anchored TLR4 (Mazgaeen and Gurung 2020). It follows that a second, indirect effect of LPS on the mouse is through its provocation of increased ERV transcription intracellularly. ERV-origin RNA, cDNA and/or protein would then be recognized by a cytoplasmic PRR. RIG-I was one associated with ERV transcription in this study. Kong et al. reported that LPS stimulated expression of Rigi in a mouse macrophages but did not investigate ERVs for an intermediary function in this phenomenon (Kong et al. 2009). As was demonstrated for LINE type retrotransposons in human fibroblasts, intracellular PRR signaling can trigger a type 1 interferon response (De Cecco et al. 2019). The combination of these two signaling events, i.e. one through surface TLR4 by LPS itself and another through intracellular PPR(s) by to-be-defined ERV products, manifested in mice and rats as a response profile that had features of both a response to a virus with type 1 interferon and ISGs and a response to a bacterial PAMP like LPS with acute phase reactants such as calprotectin and serum amyloid.

This or a similar phenomenon has been observed under other circumstances. In humans there was heightened transcription of retrotransposons in patients with septic shock (Mommert et al. 2020), as well as in peripheral blood mononuclear cells from human subjects experimentally injected with LPS (Pisano et al. 2020). Bacteria like Staphylococcus epidermidis that express TLR2 agonists, such as lipoteichoic acid, promoted expression of ERVs, which in turn modulated host immune responses (Lima-Junior et al. 2021). A synthetic analog of a B. burgdorferi lipoprotein activated human monocytic cells and promoted replication of the latent HIV virus in cells that were persistently infected (Norgard et al. 1996).

P. leucopus does not fit with this model. Instead of the prominent interferon-gamma response observed in mice and rats, there were prominent responses of interleukin-1 beta and genes associated with neutrophil activation. Instead of the much heightened expression of ISGs, like Mx2 and Isg15, in mice treated with LPS, the deermice under the same condition had a more subdued ISG transcription profile. Instead of increased expression of ERV Env protein sequences in blood of mice and rats treated with LPS, there was decreased transcription of the homologous MLV Env gene in like-treated P. leucopus.

Our first annotation of the P. leucopus genome did not identify the gene for zinc finger protein 809 (Zfp809) (A. Long et al. 2019), a KRAB domain protein that initiates ERV silencing (Wolf et al. 2015). But subsequent rounds of annotation confirmed genes and transcripts for Zfp809 in both P. leucopus (XP_028728362) and P. maniculatus (XP_006982432). In neither deermice nor mice nor rats was Zfp809 among the up- or down-regulated DEGs with LPS (Dryad Table D1).

Reducing the differences between P. leucopus and the murids M. musculus and R. norvegicus to a single all-embracing attribute may be fruitless. But from a perspective that also takes in the 2-3x longer life span of the white-footed deermouse compared to the house mouse and the capacity of P. leucopus to serve as disease agent reservoir while maintaining if not increasing its distribution (Moscarella et al. 2019), the feature that seems to best distinguish the deermouse from either the mouse or rat is its predominantly anti-inflammatory quality. The presentation of this trait likely has a complex, polygenic basis, with environmental (including microbiota) and epigenetic influences. An individual’s placement is on a spectrum or, more likely, a landscape rather than in one or another binary or Mendelian category.

One argument against a purely anti-inflammatory characterization is the greater neutrophil numbers and activity in P. leucopus compared to M. musculus in the LPS experiment. The neutrophil activation, migration, and phagocytosis would be appropriate early defenses against a pyogenic pathogen. But if not contained, they bring local and systemic risks for the host. This damage would not likely be from nitric oxide and reactive nitrogen species, given the minimal Nos2 transcription. But deermice showed heightened expression of proteases, such as Mmp8, enzymes for reactive oxygen species, such as NADPH oxidase 1 (Nox1), and facilitators of neutrophil extracellular traps, such as PAD4 (Padi4) (Table 2). We had previously identified possible mitigators, such as the protease inhibitor Slpi and superoxide dismutase 2 (Balderrama-Gutierrez et al. 2021). These findings were replicated here. The topic of neutrophil activation and these and other possible counters is considered in more detail elsewhere.

An anti-inflammatory disposition but at what cost?

An assignment of infection tolerance to a host and pathogen pairing assumes sufficient immunity against the microbe to keep it in check if elimination fails. P. leucopus, P. maniculatus, and likely some other Peromyscus species are in this sense “immunocompetent” with respect to the microbes they host and with which they may have a long history (Hoen et al. 2009). Yet, has this balance of resistance and tolerance for certain host-associated microbes been achieved in a trade-off that entails vulnerabilities to other types of agents?

The selection of LPS as the experimental model meant to cover this contingency, at least for the common denominator of acute inflammation many types of infections elicit. But LPS studies revealed potential weaknesses of P. leucopus that some pathogens might exploit. One of these is the low expression of inducible nitric oxide. Although Nos2 gene knockouts in M. musculus had lower LPS-induced mortality than their wild-type counterparts, the mutants were more susceptible to the protozoan Leishmania major and the facultative intracellular bacterium Listeria monocytogenes (Wei et al. 1995; MacMicking et al. 1995). While there are no known studies of either of these pathogens in P. leucopus, the related species P. yucatanicus is the main reservoir for Leishmania mexicana in Mexico (Chable-Santos et al. 1995). Compared with M. musculus, which suffer a high fatality rate from experimental infections with L. mexicana, P. yucatanicus infections are commonly asymptomatic (Loría-Cervera et al. 2018).

Given the restrained interferon and ISG response shown by P. leucopus, another plausible vulnerability would be viral infections. But other studies indicate that neither RNA nor DNA viruses pose an inordinately high risk for Peromyscus. Both tolerance of and resistance to the tickborne encephalitis flavivirus Powassan virus by P. leucopus were demonstrated in an experimental model in which mice, by contrast, were severely affected (Mlera et al. 2017). P. maniculatus and P. leucopus have been successfully infected with the SARS-CoV-2 virus by the respiratory route, but the infected animals displayed only mild pathology, manifested little if any disability, and recovered within a few days (Fagre et al. 2021; Griffin et al. 2021; Milovic et al. 2023). Among natural populations and in the laboratory P. maniculatus is noted for its tolerance of hantavirus, which commonly is fatal for infected humans (Childs et al. 1994; Botten et al. 2000). P. maniculatus was permissive of infection with monkeypox virus, but the infection was mild and transient (Deschambault et al. 2023).

Implications for Lyme disease and other zoonoses

Our studies of P. leucopus began with a natural population and documented a >80% prevalence of infection and high incidence of re-infections by B. burgdorferi in the area’s white-footed deermouse, the most abundant mammal there (Bunikis et al. 2004). This was a Lyme disease endemic area (Diuk-Wasser et al. 2012), where residents frequently presented for medical care for a variety of clinical manifestations, from mild to serious, of B. burgdorferi infection (Steere et al. 1986). Subclinical infections in humans occur, but most of those who become infected have a definable illness (Steere et al. 1998). The localized or systemic presence of the microbe is a necessary condition for Lyme disease, but the majority of the symptoms and signs are attributable to inflammation elicited by the organism’s presence and not from virulence properties per se or the hijacking of host cells (Coburn et al. 2021). Since humans are transmission dead-ends for B. burgdorferi and many other zoonotic agents in their life cycles, it is not surprising that human infections are generally more debilitating if not fatal than what adapted natural hosts experience.

It is in the space between the asymptomatic natural host and symptomatic inadvertent host where there may be insights with basic and translational application. With this goal, we consider the ways the results inform studies of the pathogenesis of Lyme disease, where “disease” includes lingering disorders akin to “long Covid” (Nathan 2022), and where “pathogenesis” includes both microbial and host contributions. Plausibly-germane deermouse-mouse differences identified in our studies to date are summarized in Figure 10. Two are further highlighted here.

Summary of distinguishing features of transcriptional responses in the blood between P. leucopus and M. musculus 4 h after treatment with LPS. There is semi-quantitative representation of relative transcription of selected coding sequences or ratios of transcription for selected pairs of genes in the blood.

The first is macrophage polarization (Murray 2017). By the criteria summarized above, the response to LPS by P. leucopus is consistent with the alternatively-activated or M2 type, rather than the expected classical or M1 type. But it was not only LPS-treated deermice that had this attribute, the blood of untreated animals also displayed M2 type polarization features. This included a comparatively high Arg1 expression level and a Akt1/Akt2 ratio of less than 1 at baseline in the study. This suggests that studies of other mammals, including humans, need not administer LPS or other TLR agonist to assess a blood specimen by selected RT-qPCR or equivalent assay for a disposition toward M1 or M2-type polarization. This reading could serve as a prognostic indicator of the inflammatory response to infection with B. burgdorferi or other pathogen and the long-term outcome.

The second difference we highlight is the activation of ERV transcription that was prominent in the LPS-treated mice and rats but not in similarly-treated deermice. A paradoxical enlistment of antiviral defenses, including type 1 and type 2 interferons, for an infection with an extracellular bacterium, like B. burgdorferi, may bring about more harm than benefit, especially if the resultant inflammation persists after antibiotic therapy. There are various ways to assess ERV activation in the blood, including assays for RNA, protein, and reverse transcriptase activity. A xenotropic MLV-related retrovirus has been discounted as a cause of chronic fatigue syndrome (McClure and Kaye 2010). However, production of whole virions need not occur for there to be PRR signaling in response to cytoplasmic Env protein, single stranded RNA, or cDNA (Russ and Iordanskiy 2023).

Methods

Animals

The study was carried out in accordance with the Guide for the Care and Use of Laboratory Animals: Eighth Edition of the National Academy of Sciences (https://nap.nationalacademies.org/read/12910). The protocols AUP-18-020 and AUP-21-007 were approved by the Institutional Animal Care and Use Committee of the University of California Irvine.

Peromyscus leucopus, here also referred to simply as “deermice”, were of the outbred LL stock, which originated with 38 animals captured near Linville, NC and thereafter comprised a closed colony without sib-sib matings at the Peromyscus Genetic Stock Center at the University of South Carolina (Joyner et al. 1998). The LL stock animals for this study were bred and raised at the vivarium of University of California Irvine, an AAALAC approved facility. Outbred Mus musculus breed CD-1, specifically Crl:CD1(ICR) IGS, and here also referred to as “mice”, were obtained from Charles River. Fischer F344 strain inbred Rattus norvegicus, specifically F344/NHsd, and here also referred to as “rats”, were obtained from Charles River.

For the combined P. leucopus-M. musculus experiment the 20 P. leucopus were of a mean (95% confidence interval) 158 (156-159) days of age and had a mean 21 (19-22) g body mass. The 20 M. musculus were all 149 days of age and had a mean body mass of 47 (43-50) g. The ratio of average male to average female body mass was 1.04 for P. leucopus and 1.03 for M. musculus. The 6 female P. leucopus for the 12 h duration experiment were of a mean 401 (266-535) days of age and had mean body mass of 20 (17-23) g. The 16 adult 10-12 week old female R. norvegicus had a mean 139 (137-141) g body mass.

Animals were housed in Techniplast-ventilated cages in vivarium rooms with a 16 h-8 h light-dark cycle, an ambient temperature of 22 °C, and on ad libitum water and a diet of 2020X Teklad global soy protein-free extruded rodent chow with 6% fat content (Envigo, Placentia, CA).

For injections the rodents with anesthetized with inhaled isoflurane. The rodents were euthanized by carbon dioxide overdose and intracardiac exsanguination at the termination of the experiment. No animals died or became moribund before the 4 hour or 12 h termination time points.

LPS experiments

P. leucopus and M. musculus combined experiment. All treatments were administered in the morning of a single day. At 15 min intervals and alternating between species, sex, and treatments, animals were intraperitoneally (ip) injected 50 µl volumes of either ion-exchange chromatography purified Escherichia coli O111:B4 LPS (Sigma-Aldrich L3024, batch 0000109071; <1% protein and <1% RNA) in a dose of 10 µg per g body mass or the diluent alone: sterile-filtered, endotoxin-tested 0.9% sodium chloride (Sigma-Aldrich S8776). The animals were visually monitored in separate cages continuously for the duration of the experiment. We recorded whether there was reduced activity by criterion of huddling with little or movement for > 5 min, ruffled fur or piloerection), or rapid respiration rate or tachypnea. At 4.0 h time after injection animals were euthanized as described above, and sterile dissection was carried out immediately.

Lower dose and longer duration experiment. In an experiment with 6 P. leucopus, the animals were administered the same single dose of LPS but at 1.0 µg/g and the same control solution. The animals were euthanized 12 h after the injection the following day.

Rat LPS experiment. The same experimental design was used for the rats as for the combined deermice-mice experiment, with the exception that the formulation of the E. coli O111:B4 LPS was “cell culture grade” (Sigma-Aldrich L4391), and the groups were sterile saline alone (n = 5), 5 µg LPS per g body mass (n = 6), or 20 µg LPS per g (n = 5).

Hematology and plasma analyte assays

For the combined P. leucopus-M. musculus experiment automated complete blood counts with differentials were performed at Antech Diagnostics, Fountain Valley, CA on a Siemens ADVIA 2120i with Autoslide hematology instrument with manual review of blood smears by a veterinary pathologist. For the 12 h duration P. leucopus experiment hematologic parameters were analyzed on an ABCVet Hemalyzer automated cell counter instrument at U.C. Irvine. For the rat experiment complete blood counts with differentials were performed at the Comparative Pathology Laboratory of the University of California Davis. Multiplex bead-based cytokine protein assay of the plasma of the rats was performed at Charles River Laboratories using selected options of the Millipore MILLIPLEX MAP rat cytokine/chemokine panel.

RNA extraction of blood

After the chest cavity was exposed, cardiac puncture was performed through a 25 gauge needle into a sterile 1 ml polypropylene syringe. After the needle was removed, the blood was expelled into Becton-Dickinson K2E Microtainer Tubes, which contained potassium EDTA. Anticoagulated blood was split into a sample that was placed on ice for same-day delivery to the veterinary hematology laboratory and a sample intended for RNA extraction which was transferred to an Invitrogen RiboPure tube with DNA/RNA Later and this suspension was stored at -20 °C. RNA was isolated using the Invitrogen Mouse RiboPure-Blood RNA Isolation Kit]. RNA concentration was determined on a NanoDrop microvolume spectrophotometer (ThermoFisher) and quality was assessed on an Agilent Bioanalyzer 2100.

RNA-seq

For the P. leucopus, M. musculus, and R. norvegicus RNA extracts of whole blood production of cDNA libraries was with the Ilumina TruSeq Stranded mRNA kit. After normalization and multiplexing, the libraries were sequenced at the University of California Irvine’s Genomics Research and Technology Hub on a Illumina NovaSeq 6000 instrument with paired-end chemistry and 150 cycles to achieve ∼100 million reads per sample for the combined P. leucopus-M. musculus experiment. The method for producing cDNA libraries was used for the R. norvegicus RNA but these were sequenced on a Illumina HiSeq 4000 instrument with paired-end chemistry and 100 cycles in the same facility. The quality of sequencing reads was analyzed using FastQC (Babraham Bioinformatics). The reads were trimmed of low-quality reads (Phred score of <15) and adapter sequences, and corrected for poor-quality bases using Trimmomatic.

For the combined species experiment the mean (95% confidence interval) number of PE150 reads per animal after trimming for quality was 1.1 (1.0-1.2) x 108 for P. leucopus and 1.1 (1.0-1.2) x 108 for M. musculus (p = 0.91). For the lower dose-longer duration experiment with P. leucopus the mean number of PE150 reads was 2.5 (2.3-2.6) x 107. For the rat experiment the mean number of PE100 reads was 2.4 (2.2-2.5) x 107.

Genome-wide differential gene expression

Batched fastq files were subjected to analysis with CLC Genomics Workbench version 23 (Qiagen). Library size normalization was done by the TMM (trimmed mean of M values) method of Robinson and Oshlack (Robinson and Oshlack 2010). Differential expression between experimental conditions was assessed with an assumption of a negative binomial distribution for expression level and a separate Generalized Linear Model for each (McCarthy, Chen, and Smyth 2012). Fold changes or differences were log2-transformed. The False Discovery Rate (FDR) with corrected p value was estimated by the method of Benjamini and Hochberg (Benjamini and Hochberg 1995). The reference genome transcript sets on GenBank were the following: GCF_004664715.2 for P. leucopus LL stock, GCF_000001635.27_GRCm39 for M. musculus C57Bl/6, and GCF_015227675.2_mRatBN7.2 for R. norvegicus. Principal Component Analysis was carried with the “PCA for RNA-Seq” module of the suite of programs; normalization was with Z-scores.

For the P. leucopus RNA-seq analysis there were 54,466 reference transcripts, of which 48,164 (88%) were mRNAs with protein coding sequences, and 6,302 were identified as non-coding RNAs (ncRNA). Of the 48,164 coding sequences, 40,247 (84%) had matching reads for at least one of the samples. The five most highly represented P. leucopus coding sequences among the matched transcripts of whole blood among treated and control animals were hemoglobin subunits alpha and beta, the calprotectin subunits S100A8 and S100A9, and ferritin heavy chain. For the M. musculus analysis there were available 130,329 reference transcripts: 92,486 (71%) mRNAs with protein coding sequences and 37,843 ncRNAs. Of the coding sequences, 59,239 (64%) were detectably transcribed in one or both groups by the same criterion. The five most highly represented coding sequences of mRNAs of identified genes for M. musculus were hemoglobin subunits alpha and beta, aminolevulinic synthase 2, ferritin light polypeptide 1, and thymosin beta. For R. norvegicus there were 99,126 reference transcripts, of which 74,742 (75%) were mRNAs. The five most highly represented coding sequences of mRNAs of identified genes for R. norvegicus were hemoglobin subunits alpha and beta, beta-2 microglobulin, ferritin heavy chain, and S100a9.

Gene Ontology term analysis

Enrichment of Gene Ontology (GO; http://geneontology.org) terms for biological processes was computed using EnrichR (https://amp.pharm.mssm.edu/Enrichr) (Chen et al. 2013). Mus musculus was selected as the closest reference for the P. leucopus and R. norvegicus data. The analysis was implemented using the default settings (p value cut-off of 0.01, minimum overlap of 3, and a minmum enrichment of 1.5) of the tools at the Metascape website (https://metascape.org) (Zhou et al. 2019). The GO terms were sorted by ascending p value. Besides the terms beginning with “GO”, other terms refer to Kegg Pathway database (https://www.kegg.jp) for “mmu..” designations, WikiPathways database (https://www.wikipathways.org) for “WP…” designations, and Reactome database (https://reactome.org) for “R-MMU…” designations.

Targeted RNA-seq

RNA-seq of selected set of protein coding sequences (CDS), which are listed below, was carried out using CLC Genomics Workbench v. 23 (Qiagen). Paired-end reads were mapped with a length fraction of 0.35 for ∼150 nt reads and 0.40 for ∼100 nt reads, a similarity fraction of 0.9, and penalties of 3 for mismatch, insertion, or deletion to the CDS of sets of corresponding orthologous mRNAs of P. leucopus, M. musculus, and R. norvegicus. Expression values were unique reads normalized for total reads across all the samples without adjustment for reference sequence length. Exceptions were the endogenous retrovirus coding sequences which differed in lengths between species. These were log10-transformed. For cross-species comparison we normalized for transcripts of a gene, Ptprc or CD45, that is a marker for both granulocytes and mononuclear cells in the blood. The purpose was to adjust for difference in white cell numbers between samples. The Results section provides further details. Following the recommendation of Hedges et al. we used the natural logarithm (ln) of ratios (Hedges, Gurevitch, and Curtis 1999). An alternative gene for adjustments for frequencies of white blood cells in the sample was mitochondrial 12S rRNA; mature erythrocytes do not have mitochondria. The coefficient of determination (R2) for log-transformed fold changes between Peromyscus and Mus with Ptprc-matched reads for normalization and the corresponding values with mitochondrial 12S rRNA-matched reads for normalization was 0.96 with a coefficient of 1.04. Because of the standard uses of Ptprc/CD45 as a signifier of white blood cells, we chose this gene for indexing over the 12S rRNA.

Target CDS (n=113): Acod1, Akt1, Akt2, Arg1, Bcl3, Camp, Ccl2, Ccl3, Ccl4, Cd14, Cd177, Cd3d, Cd4, Cd69, Cd8, Cfb, Cgas, Csf1, Csf1r, Csf2, Csf3, Csf3r, Cx3cr1, Cxcl1, Cxcl10, Cxcl2, Cxcl3, Dhx58, Fcer2, Fcgr2a, Fcgr2b, Fcgr3 (Cd16), Fgr, Fos, Fpr2, Gapdh, Gbp4, Glrx, Gzmb, Hif1a, Hk3, Hmox1, Ibsp, Icam, Ifih1 (Mda5), Ifit1, Ifng, Il10, Il12, Il18, Il1b, Il1rn, Il2ra, Il4ra, Il6, Il7r, Irf7, Isg15, Itgam (Cd11b), Jak1, Jak2, Jun, Lcn2, Lpo, Lrg, Lrrk2, Ltf, Mapk1, Mmp8, Mmp9, Mpo, MT-Co1, Mt2, Mtor, Mx2, Myc, Myd88, Ncf4, Nfkb1, Ngp, Nos2, Nox1, Nr3c1, Oas1, Olfm4, Padi4, Pbib, Pkm, Ptx, Retn, Rigi (Ddx58), S100a9, Saa3, Serpine1, Slc11a1 (Nramp), Slpi, Socs1, Socs3, Sod2, Stat1, Stat2, Stat4, Steap1, Sting, Tgfb, Thy1, Timp1, Tlr1, Tlr2, Tlr4, Tnf, Tnfrsf1a, and Tnfrsf9. The sources for these coding sequences were the reference genome transcript sets for P. leucopus, M. musculus, and R. norvegicus listed above under “Genome-wide differential gene expression”. If there were two or more isoforms of the mRNAs and the amino acid sequences differed, the default selection for the coding sequence was the first listed isoform. The lengths of the orthologous pairs of P. leucopus and M. musculus coding sequences were either identical or within 2% of the other. Note that Fcgr1 (Cd64) was not included in the comparison, because in P. leucopus it is a pseudogene (GenBank accession OP292976).

Reverse transcriptase-quantitative PCR assays

RT-qPCR assays and the corresponding primers for measurement of transcripts of genes for nitric oxide synthase 2 (Nos2) and glyceraldehyde 3-phosphate dehydrogenase (Gapdh) were those described previously (Balderrama-Gutierrez et al. 2021). These primers worked for M. musculus as well as P. leucopus using modified cycling conditions. For the arginase 1 transcript assays different primer sets were used for each species. The forward and reverse primer sets for the 352 bp Arg1 product for P. leucopus were 5’-TCCGCTGACAACCAACTCTG and 5’-GACAGGTGTGCCAGTAGATG, respectively. The corresponding primer pairs for a 348 bp Arg1 of M. musculus were 5’-TGTGAAGAACCCACGGTCTG and 5’-ACGTCTCGCAAGCCAATGTA. cDNA synthesis and qPCR were achieved with a Power Sybr Green RNA-to-Ct 1-Step Kit (Applied Biosystems) in 96 MicroAmp Fast Reaction Tubes using an Applied Biosystems StepOne Plus real-time PCR instrument. The initial steps for all assays were 48 °C for 30 min and 95 °C for 10 min. For Arg1 and Nos2 assays, this was followed by 40 cycles of a 2-step PCR of, first, 95°C for 15 s and then, second, annealing and extension at 60 °C for 1 min. The cycling conditions for Gapdh were 40 cycles of 95 °C for 15 s followed by 60 °C for 30 s.

Statistics

Means are presented with asymmetrical 95% confidence intervals (CI) to accommodate data that was not normally distributed. Parametric (t test) and non-parametric (Mann-Whitney) tests of significance were 2-tailed. Unless otherwise stated, the t test p value is given. Categorical variables were assessed by 2-tailed Fisher’s exact test. FDR correction of p values for multiple testing was by the Benjamini-Hochberg method (Benjamini and Hochberg 1995), as implemented in CLC Genomics Workbench (see above), or False Discovery Rate Online Calculator (https://tools.carbocation.com/FDR). Discriminant Analysis, linear regression, correlation, coefficient of determination, and General Linear Model analyses were performed with SYSTAT v. 13.1 software (Systat Software, Inc.).

Data availability

The experiments reported here are associated with National Center for Biotechnology Information (https://ncbi.nlm.nih.gov) BioProjects PRJNA975149 for the combined P. leucopus and M. musculus experiment, PRJNA874306 for the 12 h-lower dose P. leucopus experiment, and PRJNA973677 for the R. norvegicus LPS experiment. PRJNA975149 includes 40 BioSamples (SAMN35347136-SAMN35347175) and 40 sets of Sequence Read Archive (SRA) fastq files of Illumina paired-end chemistry reads 1 and 2 (SRR24733648-SRR24733687). PRJNA874306 includes 6 BioSamples (SAMN30561752-SAMN30561757) and the corresponding SRA files SRR24451178-SRR24451180, SRR24451183, SRR24451194, and SRR24451195. PRJNA973677 includes 16 BioSamples (SAMN35351370-SAMN35351385) and the corresponding SRA files SRR24731678-SRR24731693. The datasets of the differentially-expressed gene analysis of RNA-seq for the 4 h LPS experiments for P. leucopus, M. musculus, and R. norvegicus are available under the title “In vivo differentially-expressed genes in Peromyscus leucopus, Mus musculus, and Rattus norvegicus blood in response to LPS” as Table D1 at the public data repository Dryad (https://doi.org/10.7280/D1470Z).

Supporting information

eLife Assessment

This is an important study that tries to shed light on why the deer mouse is host to many diverse pathogens. The results are convincing and rely on state of the art transcriptomic analysis. The findings will be of interest to the biologists, ecologists and infectious disease researchers.

Reviewer #1 (Public Review):

Summary: A well-executed series of experiments that will likely be of immense interest to (a) vector-borne disease researchers and (b) gram-negative sepsis/bacteremia researchers. The study uses comparative transcriptomics to begin probing what makes Peromyscus leucopus a unique host for numerous pathogens. Most issues with the paper are trivial, relating to descriptions of statistical cutoffs. While the paper does not provide mechanistic insight into how P. leucopus restrains its immune response to LPS or other microbial invaders, it is likely that this paper will be frequently consulted by researchers trying to understand that phenomenon.

Strengths:

o Use of outbred M. musculus is a commendable choice for the studies here.

o Excellent decision by the authors to use their published dataset (with appropriate statistical normalization) to improve their statistical power to examine sex-biased gene expression. Is it possible to go one step further and briefly incorporate their prior BALB/c data to see how the BALB/c compare to the outbred mice. This could perhaps be just a PCA plot to see if they cluster with the outbred mice and/or Peromyscus, or are separate.

o The correlations and ratios used to try to understand immune cell dynamics are clever and likely reflect interesting biology, but caution should be used when interpreting these indirect measures. As there are no tools for cell separation in P. leucopus, the authors should continue to include these data to stimulate ideas in the field, but readers should understand the "conclusions" are hypotheses due to the nature of the bulk RNAseq.

Weaknesses:

o Supplemental Table 1 only lists genes that passed the authors statistical thresholds. The full list of genes detected in their analysis should be included with read counts, statistics, etc. as supplemental information

o While P. leucopus is a critical reservoir for B. burgdorferi, caution should be taken in directly connecting the data presented here and the Lyme disease spirochete. While it's possible that P. leucopus have a universal mechanism for limiting inflammation in response to PAMPs, B. burgdorferi lack LPS and so it is also possible the mechanisms that enable LPS tolerance and B. burgdorferi tolerance may be highly divergent.

o Statistical significance is binary and p-values should not be used as the primary comparator of groups (e.g. once a p-value crosses the deigned threshold for significance, the magnitude of that p-value no longer provides biological information). For instance, in comparing GO-terms, the reason for using of high p-value cutoffs ("None of these were up-regulated gene GO terms with p values < 1011 for M. musculus.") to compare species is unclear. If the authors wish to compare effect sizes, comparing enrichment between terms that pass a cutoff would likely be the better choice. Similarly, comparing DEG expression by p-value cutoff and effect size is more meaningful than analyses based on exclusively on p-value: "Of the top 100 DEGs for each species by ascending FDR p value." Description in later figures (e.g. Figure 4) is favored.

o The ability to use of CD45 to normalize data is unclear. Authors should elaborate both on the use of the method and provide some data how the data change when they are normalized. For instance, do correlations between untreated Mus and Peromyscus gene expression improve? The authors seem to imply this should be a standard for interspecies comparison and so it would be helpful to either provide data to support that or, if applicable, use of the technique in literature should be referenced.

o Regarding the ISG data-is a possible conclusion not that Peromyscus don't upregulate the antiviral response because it's already so high in untreated rodents? It seems untreated Peromyscus have ISG expression roughly equivalent to the LPS mice for some of the genes. This could be compared more clearly if genes were displayed as bar plots/box and whisker plots rather than in scatter plots. It is unclear why the linear regression is the key point here rather than normalized differences in expression.

o Some sections of the discussion are under supported:

 The claim that low inflammation contributes to increased lifespan is stated both in the introduction and discussion. Is there justification to support this? Do aged pathogen-free mice show more inflammation than aged Peromyscus?

 The claim that reduced Peromyscus responsiveness could lead to increased susceptibility to infection is prominently proposed but not supported by any of the literature cited.

 References to B. burgdorferi, which do not have LPS, in the discussion need to ensure that the reader understands this and the potential that responses could be very different.

Reviewer #2 (Public Review):

Milovic, Duong, and Barbour investigate the inflammatory response of three species of small mammals (P. leucopus, M. musculus, and R. norvegicus) to endotoxin lipopolysaccharide (LPS) injection via genome-wide transcriptomics from blood samples. Understanding the inflammation response of P. leucopus is of importance as they are a reservoir for several pathogens. The study is a thorough, controlled, well researched analysis that will be valuable for designing and interpreting future studies. The authors discuss the limitations of the data and the potential directions. Clearly P. leucopus respond differently to the LPS exposure which is very interesting and opens the door for numerous other comparative studies.

The conclusions of the manuscript are thoughtful and mostly supported by the data, but there are a couple of points for clarification.

1. How were the number of animals for each experiment selected? Was a power analysis conducted?

2. The authors conducted a cursory evaluation of sex differences of P. leucopus and reported no difference in response except for Il6 and Il10 expression being higher in the males than the females in the exposed group. The data was not presented in the manuscript. Nor was sex considered for the other two species. A further discussion of the role that sex could play and future studies would be appreciated.

3. The ratio of Nos2 and Arg1 copies for LPS treated and control P. leucopus and M.musculus in Table 3 show that in P. leucopus there is not a significant difference but in M.musculus there is an increase in Nos2 copies with LPS treatment. The authors then used a targeted RNA-seq analysis to show that in P. leucopus the number of Arg1 reads after LPS treatment is significantly higher than the controls. These results are over oversimplified in the text as an inverse relationship for Nos2/Arg1 in the two species.

Author Response

The following is the authors’ response to the original reviews.

Re: Revised author response for eLife-RP-RA-2023-90135 (“The white-footed deermouse, an infection-tolerant reservoir for several zoonotic agents, tempers interferon responses to endotoxin in comparison to the mouse and rat” by Milovic, Duong, and Barbour”)

The revised manuscript has taken into account all the comments and questions of the two reviewers. Our responses to each of the comments are detailed below. In brief, the modifications or additional materials for the revision each specifically address a reviewer comment. These modifcations or materials include the following….

• a more in-depth consideration of sample sizes

• a better explanation of what p values signify for a GO term analysis

• a more detailed account of the selection of the normalization procedure for cross-species targeted RNA-seq (including a new supplemental figure)

• several more box plots in supplementary materials to complement the scatterplots and linear regressions of the figures of the primary text

• provision in a public access repository of the complete data for the RNA-seq analyses as well as primary data for figures and tables as new supplementary tables

• the expansion of description of the analysis done for the revision of Borrelia hermsii infection of P. leucopus. This included a new table (Table 10 of the revision) • development of the possible relevance of finding for longevity studies by citing similarities of the findings in P. leucopus with those in the naked mole-rat

• what we think is a better assessment of differences between female and male P. leucopus for this particular study, while still keeping focus on DEGs in common for females and males. This included a new figure (Figure 4 of the revision).

• removal of reference to a “inverse” relationship between Nos2 and Arg1 while still retaining ratios of informative value

We note that in the interval between uploading the original bioRxiv preprint and now we learned of the paper of Gozashti, Feschotte, and Hoekstra (reference 32), which supports our conception of the important place of endogenous retroviruses in the biology and ecology of deermice. This is the only addition or modification that was not a direct response to a reviewer comment or question, but it was germane to one of Reviewer #1’s comments (“Regarding..”).

Reviewer #1:

Supplemental Table 1 only lists genes that passed the authors statistical thresholds. The full list of genes detected in their analysis should be included with read counts, statistics, etc. as supplemental information.

We agree that provision of the entire lists of reference transcripts and the RNA-seq results for each of the 40 animals is merited. These datasets are too large for what the journal’s supplementary materials resource was intended for, so we have deposited them at the Dryad public access repository.

While P. leucopus is a critical reservoir for B. burgdorferi, caution should be taken in directly connecting the data presented here and the Lyme disease spirochete. While it's possible that P. leucopus have a universal mechanism for limiting inflammation in response to PAMPs, B. burgdorferi lack LPS and so it is also possible the mechanisms that enable LPS tolerance and B. burgdorferi tolerance may be highly divergent.

The impetus for the study was the phenomenon of tolerance of infection of P. leucopus by a number of different kinds of pathogens, not just B. burgdorferi. We take the reviewer’s point, though. Certainly, the white-footed deermouse is probably most notable at-large for its role as a reservoir for the Lyme disease agent. We doubt that the species responses to LPS and to the principal agonists of B. burgdorferi are “highly divergent”, though. Other than the TLR itself-TLR4 for LPS vs the heterodimer TLR2/TLR1 for the lipoproteins of these spirochetes--the downstream signaling is generally similar for amounts comparable in their agonist potency.

We had thought that we had addressed this distinction for B. burgdorferi and other Borreliaceae members by referring to the earlier study. But we agree with the reviewer that what was provided on this point was insufficient in the context of the present work. Accordingly, for the revision we have added a new analysis of the data on experimental infection of P. leucopus with Borrelia hermsii, which lacks LPS and for which the TLR agonists eliciting inflammation are lipoproteins. We do this in a format (new Table 6) that aids comparison with the LPS experimental data elsewhere in the article. As the manuscript references, B. burgdorferi infection of P. leucopus elicits comparatively little inflammation in blood even at the height of infection. While this phenomenon with the Lyme disease agent was part of the rationale driving these studies, the better comparison with LPS was 5 days into B. hermsii infection when the animals are spirochetemic.

Statistical significance is binary and p-values should not be used as the primary comparator of groups (e.g. once a p-value crosses the deigned threshold for significance, the magnitude of that p-value no longer provides biological information). For instance, in comparing GO-terms, the reason for using of high p-value cutoffs ("None of these were up-regulated gene GO terms with p values < 1011 for M. musculus.") to compare species is unclear. If the authors wish to compare effect sizes, comparing enrichment between terms that pass a cutoff would likely be the better choice. Similarly, comparing DEG expression by p-value cutoff and effect size is more meaningful than analyses based on exclusively on p-value: "Of the top 100 DEGs for each species by ascending FDR p value." Description in later figures (e.g. Figure 4) is favored.

Effect sizes--in this case, fold-changes--were taken into account for GO term analysis and were specified in the settings that are described. So, any gene that was “counted” for consideration for a particular GO term would have passed that threshold and with a falsediscovery corrected p value of a specified minimum. There is no further scoring of the “hit” based upon the magnitude of the p value beyond that point. It is, as the reviewer writes, binary at that point. We are in agreement on those principles.

As we understand the comment above, though, the p-values referred to are in regard to the GO term analysis itself. The objective was discovery followed by inference. The situation was more like a genome-wide association study (GWAS) study. This is not strictly speaking a hypothesis test, because there was no stated hypothesis ahead of time or one driving the design. The “p value” for something like GO term analysis or GWAS provides an estimate of the strength of the association. It is not binary in that sense. The lower the p value, the greater confidence about the association. In a GWAS of a human population an association of a trait with a particular SNP or indel is usually not taken seriously unless the p value is less than 10^-7 or 10^-8. In the case of GO terms, the p value approximates (but is not equivalent to) the number of genes that are differentially expressed that belong to a GO cluster out of the total number of genes that define that cluster. The higher the proportion of the genes in the cluster that are associated with a treatment (LPS vs. saline), the lower the p value. Thus, it provides information beyond the point at which it would be rightly deemed of little additional value in many hypothesis testing circumstances.

That said, we agree that the original manuscript could have been clearer on this point and have for the revision expanded the description of the GO term analysis in the Methods, including some explanation for a reader on what the p value signifies here. We also refrain from specifying a certain p value for special attention and merely list 20 by ascending p value.

The ability to use of CD45 to normalize data is unclear. Authors should elaborate both on the use of the method and provide some data how the data change when they are normalized. For instance, do correlations between untreated Mus and Peromyscus gene expression improve? The authors seem to imply this should be a standard for interspecies comparison and so it would be helpful to either provide data to support that or, if applicable, use of the technique in literature should be referenced.

The reviewer brings up an important point that we considered addressing in more depth for the original manuscript but in the end deferred to considerations about length and left it out.

But we are glad to address this here, as well as in the revised manuscript.

We did not intend to imply either that this particular normalization approach had been done before by others or that it “should” be a standard. We are not aware of another report on this, and it would be up to others whether it would be useful or not for them. We made no claim about its utility in another model or circumstance. The challenge before us was to do a comparative analysis of transcription in the blood not just for animals of one species under different conditions but animals of two different genera under different conditions. A notable difference between the animals was in their white blood cell counts, as this study documents. White cells would be the source of a majority of transcripts of potential relevance here, but there would also be mRNA for globins, from reticulocytes, from megakaryocytes, and likely cell-free RNA with origins in various tissues. If the white cell numbers differed, but the non-white cell sources of RNA did not, then there could be unacknowledged biases.

It would be like comparing two different kinds of tissues and assuming them to be the same in the types and numbers of cells they contained. Four hours after a dose of LPS the liver cells (or brain cells) would differ in their transcriptional profiles from untreated the livers (or brains) of untreated animals for sure, but there would not be much if any change in the numbers of different kinds of cells in the liver (or brain) within 4 hours. The blood can change a lot in composition within that time frame under these same conditions. Some sort of accounting for differing white cell numbers in the blood in different outbred animals of two species seemed to be called for.

The normalization that was done for the genome-wide analysis was not based on a particular transcript, but instead was based on the total number of reads, the lengths of the reference transcripts, and the distributions of reads matching to the tens of thousands of references for each sample. This was done according to what are standard procedures by now for bulk RNAseq analyses. Because the reference transcript sets for P. leucopus and M. musculus differed in their numbers and completeness of annotation, we did not attempt any cross-species comparison for the same set of genes at that point. That would not be possible because they were not entirely commensurate.

The GO term analysis of those results provided the leads for the more targeted approach, which was roughly analogous to RT-qPCR. For a targeted assay of this sort, it is common to have a “housekeeping gene” or some other presumably stably transcribed gene for normalization. A commonly used one is Gapdh, but we had previously found that Gapdh was a DEG itself in the blood in P. leucopus and M. musculus at the four hour mark after LPS. The aim was to provide for some adjustment so datasets for blood samples differing in white blood cell counts could be compared. Two options were the 12S ribosomal RNA of the mitochondria, which would be in white cells but not mature erythrocytes, and CD45, which has served an approximately similar function for flow cytometry of the blood. As described in what has been added for the revision and the supplementary materials, we compared these different approaches to normalization. Ptprc and 12S rRNA were effectively interchangeable as the denominator with identifying DEGs of P. leucopus and M. musculus and cross-species comparisons.

Regarding the ISG data-is a possible conclusion not that Peromyscus don't upregulate the antiviral response because it's already so high in untreated rodents? It seems untreated Peromyscus have ISG expression roughly equivalent to the LPS mice for some of the genes. This could be compared more clearly if genes were displayed as bar plots/box and whisker plots rather than in scatter plots. It is unclear why the linear regression is the key point here rather than normalized differences in expression.

In answer to the question: yes, that is possible. In the interval between uploading of the manuscript and this revision, we became aware of a study by Gozashti and Hoekstra published this year in Molecular Biology and Evolution (reference 32) and reporting on the “massive invasion” of endogenous retroviruses in P. maniculatus and the defenses deployed in response to achieve silencing. We cite this work and discuss it, including related findings for P. leucopus, in the revision.

We had originally intended to include box plots as well as scatterplots with regressions for the data, but thought it would be too much and possibly considered redundant. But with this encouragement from the reviewer we provide additional box plots in supplementary materials for the revision.

Some sections of the discussion are under supported:

The claim that low inflammation contributes to increased lifespan is stated both in the introduction and discussion. Is there justification to support this? Do aged pathogen-free mice show more inflammation than aged Peromyscus?

We respectively point out that there was not a claim of this sort. We stated a fact about P. leucopus’ longevity. We made no statement connecting longevity and inflammation beyond the suggestion in the introduction that the explanation(s) for infection tolerance might have some bearing for studies on determinants of life span.

But the reviewer’s comment prompted further consideration of this aspect of Peromyscus biology. This led eventually to the literature on the naked mole-rat, which seems to be the rodent with the longest known life span and the subject of considerable study. The discussion section of the revision has an added paragraph on some of the similarities of P. leucopus and the naked mole-rat in terms of neutrophils, expression of nitric oxide synthase 2 in response to LPS, and type 1 interferon responses. While this is far from decisive, it does serve to connect some of the dots and, hopefully, is considered at least partially responsive to the reviewer’s question.

The claim that reduced Peromyscus responsiveness could lead to increased susceptibility to infection is prominently proposed but not supported by any of the literature cited.

There was not this claim. In fact, it was framed as a question, not a statement. Nevertheless, we think we understand what the comment is getting at and acknowledge in the revision that there may be unexamined circumstances in which P. leucopus may be more vulnerable.

References to B. burgdorferi, which do not have LPS, in the discussion need to ensure that the reader understands this and the potential that responses could be very different.

We think we addressed this comment in a response above.

Reviewer #2:

1. How were the number of animals for each experiment selected? Was a power analysis conducted?

A power analysis of any meaning for bulk RNA-seq with tens of thousands of reference transcripts, each with their own variance, and a comparison of animals of two different genera is not straight forward. Furthermore, a specific hypothesis was not being tested. This was a broad, forward screen. But the question about sample sizes is one that deserves more attention than the original manuscript provided. This now provided in added text in two places in Methods ( “RNA-seq” and “Genome-wide different gene expression”) in the revision.

1. The authors conducted a cursory evaluation of sex differences of P. leucopus and reported no difference in response except for Il6 and Il10 expression being higher in the males than the females in the exposed group. The data was not presented in the manuscript. Nor was sex considered for the other two species. A further discussion of the role that sex could play and future studies would be appreciated.

We agree that the limited analysis of sex differences and the undocumented remark about Il6 and Il10 expression in females and males warranted correction. For the revision we removed that analysis of targeted RNA-seq of P. leucopus from the two different studies. For this study we were looking for differences that applied to both species. This was the reason that there were equal numbers of females and males in the samples. We agree that further investigation of differences between sexes in their responses is of interest but is probably best left for “future studies”.

But in revision we do not entirely ignore the question of sex of the animal and provide an additional analysis of the bulk RNA-seq for P. leucopus with regard to differences between females and males. This basically demonstarted an overall commensurability between sexes, at least for the purposes of the GO term analysis and subsequent targeted RNA-seq, but did reveal some exceptions that are candidate genes for those future studies.

In the revision, we also add for the discussion and its “study limitations” section a disclaimer about possibly missing sex associated differences because the groups were mixed sexes.

1. The ratio of Nos2 and Arg1 copies for LPS treated and control P. leucopus and M.musculus in Table 3 show that in P. leucopus there is not a significant difference but in M.musculus there is an increase in Nos2 copies with LPS treatment. The authors then used a targeted RNA-seq analysis to show that in P. leucopus the number of Arg1 reads after LPS treatment is significantly higher than the controls. These results are over oversimplified in the text as an inverse relationship for Nos2/Arg1 in the two species.

We agree. In addition to providing box plots for Arg1 and Nos2, as suggested by Reviewer #1, we also replaced “ratio” in commenting on Arg1 and Nos2, with “differences in Nos2 and Arg1 expresssion” replacing “ratio of Nos2 to Arg1 expression” at one place. At another place we have removed “inverse” with regard to Nos2 and Arg1. But we respectfully decline to remove Nos2/Arg1 from Figure 5 (now Figure 6) or inclusion of Nos2/Arg1 ratios elsewhere. According to our understanding there need not be an inverse relationship for a ratio to have informative value.

Recommendations For the Authors

We thank the two reviewers for their constructive recommendations and suggestions, in some case pointing out errors we totally missed. For the great majority, the recommendations were followed. Where we decline or disagree we explain this in the response.

Reviewer #1 (Recommendations For The Authors):

• How was the FDR < 0.003 cutoff chosen for DEG? All cutoffs are arbitrary but there should be some justification.

We agree and have provided the rationale at that point in the paper (before Figure 3) in R2: "For GO term analysis the absolute fold-change criterion was ≥ 2. Because of the ~3-fold greater number of transcripts for the M. musculus reference set than the P. leucopus reference set, application of the same false-discovery rate (FDR) threshold for both datasets would favor the labeling of transcripts as DEGs in P. leucopus. Accordingly, the FDR p values were arbitrarily set at <5 x 10-5 for P. leucopus and <3 x 10-3 for M. musculus to provide approximately the same number of DEGs for P. leucopus (1154 DEGs) and M. musculus (1266 DEGs) for the GO term comparison."

• It would be helpful to include a figure demonstrating the correlation between CD45 and WBC ("Pearson's continuous and Spearman's ranked correlations between log-transformed total white blood cell counts and normalized reads for Ptprc across 40 animals representing both species, sexes, and treatments were 0.40 (p = 0.01) and 0.34 (p = 0.03), respectively.")

In both the first version of the revision (R1) and in R2 we provide a fuller explanation of the choice of CD45 (Ptprc) for normalization as detailed in the response to Reviewer #1's public comment. In the revision only Pearson's correlation and p value is given. We did not think another figure was justified after there was additional space devoted to this in both R1 and R2.

• Unclear what the following paragraph is referring to-is this from the previous paper? Was this experiment introduced somewhere? "Low transcription of Nos2 and high transcription of Arg1 both in controls and LPS-treated P. leucopus was also observed in the experiment where the dose of LPS was 1 µg/g body mass instead of 10 µg/g and the interval between injection and assessment was 12 h instead of 4 h (Table 4)."

This experiment is described in the Methods in the original and subsequent versions, but we agree that it is not clear whether it was from present study or previous one. Here is the revised text for R2: "Low transcription of Nos2 in both in controls and LPS-treated P. leucopus and an increase in Arg1 with LPS was also observed in another experiment for the present study where the dose of LPS was 1 µg/g body mass instead of 10 µg/g and the interval between injection and assessment was 12 h instead of 4 h (Table 4)."

• Regarding the differences in IFNy between outbred and BALB/c mice-are there any other RNA-seq datasets you can mine where other inbred mice (B/6, C3H, etc) have been injected with LPS and probed roughly the same amount of time later? Do they look like BALB/c or the outbreds?

In both the original and R1 and R2 we cite two papers on the difference of BALB/c mice. While this is of interest for follow-up in the future, we did not think additional content on a subject that mainly pertains to M. musculus was warranted here, where the main focus is Peromyscus.

• Figure 8 and its legend are difficult to follow. The top half of the figure is not well explained and it's unclear what species this is. Decreased use of abbreviations would help. Consider marking each R2 value as Mus or Peromyscus (As done in Fig 9). There are some typographical errors in the legend ("gree," incomplete sentence missing the words LPS or treatment AND Mus: "Co-variation between transcripts for selected PRRs (yellow) and ISGs (gree) in the blood of P. leucopus (P) or (M) with (L") or without (C)."

This is now Figure 9 in both R1 and R2. We revised it for R1 to include references to the box plots in supplementary materials, but agree with Reviewer #1's recommendation to correct the typos and make the legend less confusing. We did not think that further labeling of the R2 values in the scatterplots with the species names was necessary. The data points are not just colors but also different symbols, so it should be fairly easy for readers to distinguish the regression lines by species. For R2 this is the revised legend with additions in response to the recommendation underlined:

"Figure 9. Co-variation between transcripts for selected PRRs and ISGs in the blood of P. leucopus (P) or M. musculus (M) with (L) or without (C) LPS treatment. Top panel: matrix of coefficients of determination (R2) for combined P. leucopus and M. musculus data. PRRs are indicated by yellow fill and ISGs by blue fill on horizontal and vertical axes. Shades of green of the matrix cells correspond to R2 values, where cells with values less than 0.30 have white fill and those of 0.90-1.00 have deepest green fill. Bottom panels: scatter plots of log-transformed normalized Mx2 transcripts on Rigi (left), Ifih1 (center), and Gbp4 (right). The linear regression curves are for each species. For the right-lower graph the result from the General Linear Model (GLM) estimate is also given. Values for analysis are in Table S4; box plots for Gbp4, Irf7, Isg15, Mx2, and Oas1 are provided in Figure S6."

• Discussion section could benefit from editing for clarity. Examples listed: o Unclear what effect is described here "The bacterial infection experiment indicated that the observed effect in P. leucopus was not limited to a TLR4 agonist; the lipoproteins of B. hermsii are agonists for TLR2 (Salazar et al. 2009)."

Both R1 and R2 include the new section on the B. hermsii infection model. This was added in response to Reviewer #1 public comment. So the expanded consideration of this aspect should address the reviewer's recommendation for more clarity and context here. For R2 we modified the text in the discussion of R1:

"The analysis here of the B. hermsii infection experiment also indicated that the phenomenon observed in P. leucopus was not limited to a TLR4 agonist."

o Unclear what the takeaway from this paragraph is: "Reducing the differences between P. leucopus and the murids M. musculus and R. norvegicus to a single all-embracing attribute may be fruitless. But from a perspective that also takes in the 2-3x longer life span of the whitefooted deer mouse compared to the house mouse and the capacity of P. leucopus to serve as disease agent reservoir while maintaining if not increasing its distribution (Moscarella et al. 2019), the feature that seems to best distinguish the deer mouse from either the mouse or rat is its predominantly anti-inflammatory quality. The presentation of this trait likely has a complex, polygenic basis, with environmental (including microbiota) and epigenetic influences. An individual's placement is on a spectrum or, more likely, a landscape rather than in one or another binary or Mendelian category."

We agree that modification, simplication, and clarification was called for. In response to a public comment of Reviewer #1 we had changed that section, leaving out reference to longevity here. Here is the revised text in both R1 and R2:

"Reducing differences between P. leucopus and murids M. musculus and R. norvegicus to a single attribute, such as the documented inactivation of the Fcgr1 gene in P. leucopus (7), may be fruitless. But the feature that may best distinguish the deermouse from the mouse and rat is its predominantly anti-inflammatory quality. This characteristic likely has a complex, polygenic basis, with environmental (including microbiota) and epigenetic influences. An individual’s placement is on a spectrum or, more likely, a landscape rather than in one or another binary or Mendelian category."

Minor comments:

• Use of blue and red in figures as the -only- way to easily distinguish between groups is a poor choice-both in terms of how inclusivity of color-blind researchers and enabling grayscale printing. Most detrimental in Figure 2, but also slightly problematic in Figure 1. Use of color and shape (as done in other figures) is a much better alternative.

We agree. Both figures have been modified to include an additional characteristic for denoting the data point. For Figure 1 it is a black filling, and for Figure 2 it is the size of symbol in additon to the color. This should enable accurate visualization by color blind individuals and printing in gray scale. We have added definitions for the symbols within the graph itself, so there is no need to refer to the legend to interpret what they mean.

• Note the typo where it should read P leucopus: "The differences between P. musculus and M. musculus in the ratios of Nos2/Arg1 and IL12/IL10 were reported before (BalderramaGutierrez et al. 2021),"

We thank the reviewer for pointing this typo out, which also carried over to R1. It has been corrected for R2.

• Optional: Can the relationship between the ratios in figure 5 and macrophage "types" be displayed graphically alongside the graphs? It's a little challenging to go back and forth between the text and the figure to try to understand the biological implication.

We considered something like this but in the end decided that we were not yet comfortable assigning “types” in this fashion for Peromyscus.

Reviewer #2 (Recommendations For The Authors):

• Be consistent with nomenclature for your species/treatment groups in the text, figures, and tables. For example, you go back and forth between "P. leucopus" and "deermouse" in the text. And in figures you use "P," "Peromyscus", or "Pero".

In the Methods section of the original and revisions R1 and R2 we indicate that "deermouse" is synonymous with "Peromyscus leucopus" and "mouse" is synonymous with "Mus musculus" in the context of this paper. We think that some alternation in the terms relieves the text of some of its repetitiveness and that readers should not have a problem with equating one with the other. The use of "deermouse" also reinforces for readers that Peromyscus is not a mouse. With regard to the abbreviations for P. leucopus, those were used to accommodate design and space issues of the figures or tables. In all cases, the abbreviations referred to are defined in the legends of the figures. So, we respectfully decline to follow this recommendation.

• Often the sentence structure and/or word choice is irregular and makes quick/easy comprehension difficult. Several examples are:

o The third paragraph of the introduction

We agree that the first and second sentences are unclear. Here is the revision for R2:

“As a species native to North America, P. leucopus is an advantageous alternative to the Eurasian-origin house mouse for study of natural variation in populations that are readily accessible (9, 53). A disadvantage for the study of any Peromyscus species is the limited reagents and genetic tools of the sorts that are applied for mouse studies.”

o The first line after Figure 5 on page 9.

We agree. The long sentence which we think the reviewer is referring to has been in split into two sentences for R2.

“An ortholog of Ly6C (13), a protein used for typing mouse monocytes and other white cells, has not been identified in Peromyscus or other Cricetidae family members. Therefore, for this study the comparison with Cd14 is with Cd16 or Fcgr3, which deermice and other cricetines do have.”

o The sentence that starts "Our attention was drawn to..." on page 14.

We agree that the sentence was awkward and split into two sentences.

“Our attention was drawn to ERVs by finding in the genome-wide RNA-seq of LPS-treated and control rats. Two of the three highest scoring DEGs by FDR p value and fold-change were a gagpol polyprotein of a leukemia virus with 131x fold-change from controls and a mouse leukmia virus (MLV) envelope (Env) protein with 62x fold-change (Dryad Table D5).”

• For figures with multiple panels, use A), B) etc then indicate which panel you are discussing in your text. This is a very data heavy study and your readers can easily get lost.

We agree and have added pointers in the text to the panels we are referring to. But we prefer to use easily understood descriptors like “left” and “upper” over assigned letters.

• For all the figures, where are the stats from the t-tests? Why didn't you do a two-way ANOVA? Instead of multiple t-tests?

Where we are not hypothesis testing and we are able to show all the data points in box-whisker plots with distributions fully revealed, our default position is not to apply significance tests in a post hoc fashion. If a reader or other investigator wants to do this for other purposes, e.g. a meta-analysis, the data is provided in public repository for them to do this. We are not sure what the reviewer means by "multiple t-tests" for "all figures". Where we do 2-tailed t-tests for presentation of data for many genes in a table for the targeted RNA (where individual values cannot shown in the table), there is always correction for multiple testing, as indicated in Methods. The p values shown as "FDR" are after correction.

• Results paragraph "LPS experiment and hematology studies"

o List the two species for the first description to orient the reader since you eventually include rat data.

We agree that this is warranted and followed this recommendation for R2.

o Not all the mice experienced tachypnea, but the text makes it seem like 100% did.

We are not sure what the reviewer is referring to here. This is what is in the text on tachypnea: "By the experiment’s termination at 4 h, 8 of 10 M. musculus treated with LPS had tachypnea, while only one of ten LPS-treated P. leucopus displayed this sign of the sepsis state (p = 0.005)." The only other mention of "tachypnea" was in Methods.

• Figure 1: Why was the M. musculus outlier excluded? Where any other outliers excluded?

That data point for the mouse was not "excluded" from the graph. It is identified (MM17) for reference with Table 1, and there is the graph for all to see where it is. It was only excluded from the regression curve for control mice. There was no significance testing. There were no other outliers excluded.

• Figure 3: explain the colors and make the scales the same for all the panels or at least for the upregulated DEGs and the downregulated DEGs.

We have modified the legend for Figure 3 to include fuller definitions of the x-axes and a description of the color spectrum. We decline to make the x-axis scale the same for all the panels because the horizontal bars in “transcription down” panels would take up only a small fraction of the space. The x-axes are clearly defined and the colors of the bars also indicate the differences in p-values. We doubt that readers will be misled. Here is the revised legend: “Figure 3. Gene Ontology (GO) term clusters associated with up-regulated genes (upper panels) and down-regulated genes (lower panels) of P. leucopus (left panels) and M. musculus (right panels) treated with LPS in comparison with untreated controls of each species. The scale for the x-axes for the panels was determined by the highest -log10 p values in each of the 4 sets. The horizontal bar color, which ranges from white to dark brown through shades of yellow through orange in between, is a schematic representation of the -log10 p values.”

• Results paragraph "Targeted RNA seq analysis"

o In the third paragraph, an R2 of 0.75 is not close enough to 1 to call it "~1"

What the reviewer is referring to is no longer in either R1 and R2, as detailed in the authors' response to public comments.

o In the 4th paragraph, where are your stats?

We have replaced terms like "substantially" and "marginally" with simple descriptions of relationships in the graphs.

"For the LPS-treated animals there was, as expected for this selected set, higher expression of the majority genes and greater heterogeneity among P. leucopus and M. musculus animals in their responses for represented genes. In contrast to the findings with controls, Ifng and Nos2 had higher transcription in treated mice. In deermice the magnitude of difference in the transcription between controls and LPS-treated was less."

• Figure 4: The colors are hard to see, I suggest making all the up regulated reads one color, the down regulated reads a different color, and the reads that aren't different black or gray.

This is now Figure 5 in R1 and R2. The selected genes that are highlighted in the panels are denoted not only by color but also by type of symbol. We do not think that readers will have a problem telling one from another even if color blind. The purpose of this figure was to provide an overview and a visual representation with calling out of selected genes, some of which will be evaluated in more detail later. We thought that this was necessary before diving deeper into the data of Table 2. We do not think further discriminating between transcripts in the categorical way that the reviewer suggests is warranted at this point. So, we respectfully decline to follow this suggestion.

• Results paragraph " Alternatively- activated macrophages...."

o Include a brief description of Nos2 and Arg1

We have defined what enzymes these are genes for in R2.

o How do you explain the lack of a difference in P. leucopus Arg1? Your text says the RT-qPCR confirms the RNA-seq findings.

There was a difference in P. leucopus Arg1 by RT-qPCR between control and LPS treated by about 3-fold. By both RNA-seq and RT-qPCR Arg1 transcription is higher in P. leucopus than in M. musculus under both conditions. But we have modified the sentence so that does not imply more than what the data and analysis of the table reveal.

"While we could not type single cells using protein markers, we could assess relative transcription of established indicators of different white cell subpopulations in whole blood. The present study, which incorporated outbred M. musculus instead of an inbred strain, confirmed the previous finding of differences in Nos2 and Arg1 expression between M. musculus and P. leucopus (Figure 5; Table 2). Results similar to the RNA-seq findings were obtained with specific RT-qPCR assays for Nos2 and Arg1 transcripts for P. musculus and M. musculus (Table 3)."

• Figure 5: reorganize the panels to make the text description and label with letters, where are the stats?

We thought the figure (now Figure 6) was self-explanatory, but agree that further explanation in the legend was indicated. We prefer to use descriptions of locations (“upper left”) over labels, like “panel C”, which do not obviously indicate the location of the panel. Of course, if the journal’s style mandates the other format we will do so. Our response about “stats” for boxplot figures is the same as what we provided above.

• Results paragraph "Interferon-gamma and interleukin-1 beta..."

o Either add the numbers or direct the viewer to where Ifng is in Table 2. The table is very big and Ifng is all the way at the bottom!

We agree that this table is large, but we thought it better to err on the side of inclusiveness by having a single table, rather than have some genes in the main article and other results in a supplementary table. We thought that it would make it easier for reviewers and readers to find a gene of interest, but we also acknowledge the challenge to locate the genes we highlight. We follow for R2 that reviewer's recommendation to provide some guidance for readers trying to locate a featured gene by pointing relative locations. While adding a column of numbers to already complex table seems more than what is called for, we are depositing an Excel spreadsheet of the table at the Dryad repository to facilitate searching by an interested reader for a particular gene.

• Figure 6: stats? The pink and red are hard to easily distinguish from each other. I also suggest not using red and green together for color blind readers.

With regard to the box-plots and significance testing, please see response above to an earlier recommendation. We have removed an interpretative adjective (i.e. "marked") from the description of the graph. Different symbols as well as colors are used, so we do not think that this will pose a problem for readers, even those with complete red-green color blindness. For what it’s worth, with regard to the "red" and "pink" issue, according to the figure on our displays the colors of the two symbols appear to be red and purple. They are also applied to different species and different conditions for those species.

• Figure 8: In the legend it says "... PRRs (yellow) and ISGs (gree)" which is a typo, but don't you mean blue not green anyways?

See response above to Reviewer #1's recommendation. This has been corrected.

Version 2

Abstract

Abstract

The white-footed deermouse Peromyscus leucopus, a long-lived rodent, is a key reservoir for agents of several zoonoses, including Lyme disease. While persistently infected, this deermouse is without apparent disability or diminished fitness. For a model for inflammation elicited by various pathogens, the endotoxin lipopolysaccharide (LPS) was used to compare genome-wide transcription in blood by P. leucopus, Mus musculus and Rattus norvegicus and adjusted for white cell concentrations. Deermice were distinguished from the mice and rats by LPS response profiles consistent with non-classical monocytes and alternatively-activated macrophages. LPS-treated P. leucopus, in contrast to mice and rats, also displayed little transcription of interferon-gamma and lower magnitude fold-changes in type 1 interferon-stimulated genes. This was associated with comparatively reduced transcription of endogenous retrovirus sequences and cytoplasmic pattern recognition receptors in the deermice. The results reveal a mechanism for infection tolerance in this species and perhaps other animal reservoirs for agents of human disease.

Introduction

How does the white-footed deermouse Peromyscus leucopus continue to thrive while sustaining infections with disease agents it serves as reservoir for (4)? The diverse tickborne pathogens (and diseases) for humans include the extracellular bacterium Borreliella burgdorferi (Lyme disease), the intracellular bacterium Anaplasma phagocytophilum (anaplasmosis), the protozoan Babesia microti (babesiosis), and the Powassan flavivirus (viral encephalitis). Most deermice remain persistently infected but display scant inflammation in affected tissues(23, 52, 64), and without apparent consequence for fitness (83, 92).

A related question—conceivably with the same answer--is what accounts for the two-to-three fold longer life span for P. leucopus than for the house mouse, Mus musculus (48, 79)? The abundance of P. leucopus across much of North America (34, 66) and its adaptation to a variety of environments, including urban areas and toxic waste sites (11, 50, 67), indicates successful adjustment to changing landscapes and climate. Peromyscus species, including the hantavirus reservoir P. maniculatus (65), are more closely related to hamsters and voles in family Cricetidae than to mice and rats of family Muridae (15).

As a species native to North America, P. leucopus is an advantageous alternative to the Eurasian-origin house mouse for study of natural variation in populations that are readily accessible (9, 53). A disadvantage for the study of any Peromyscus species is the limited reagents and genetic tools of the sorts that are applied for mouse studies. As an alternative, we study P. leucopus with a non-reductionist approach that is comparative in design and agnostic in assumptions (3). The genome-wide expression comparison for P. leucopus is with M. musculus and, added here, the brown rat Rattus norvegicus. Given the wide range of pathogenic microbes that deermice tolerate, we use the bacterial endotoxin lipopolysaccharide (LPS) as the primary experimental treatment because the inflammation it elicits within a few hours has features common to different kinds of serious infections, not to mention severe burns and critical injuries (96).

We previously reported that a few hours after injection of LPS, P. leucopus and M. musculus had distinguishing profiles of differentially expressed genes (DEG) in the blood, spleen, and liver (3). In brief, the inflammation phenotype of deermice was consistent with an “alternatively activated” or M2-type macrophage polarization phenotype instead of the expected “classically activated” or M1-type polarization phenotype that was observed for M. musculus (69). The deermice also differed from mice in displaying evidence of greater neutrophil activation and degranulation after LPS exposure. The potentially damaging action from neutrophil proteases and reactive oxygen species appeared to be mitigated in part in P. leucopus by proteins like secretory leukocyte protease inhibitor (Slpi) and superoxide dismutase 2 (Sod2).

Here, we first address whether the heightened transcription of neutrophil-associated genes in P. leucopus is attributable to differences in numbers of white cells in the blood. To better match for genetic diversity, we substituted outbred M. musculus for the inbred BALB/c mouse of the previous study. We retained the experimental protocol of short-term responses to LPS. This main experiment was supplemented by a study of rats under the similar conditions, by an investigation of a different dose of LPS and duration of exposure in another group of deermice, and by analysis of deermice infected with a bacterium lacking LPS. The focus was on the blood of these animals, not only because the distinctions between species in their transcriptional profiles were nearly as numerous for this specimen as for spleen and liver (3), but also because for ecological and immunological studies of natural populations of Peromyscus species blood is obtainable from captured-released animals without their sacrifice.

The results inform future studies of Peromyscus species, not only with respect to microbial infections and innate immunity, but conceivably also determinants of longevity and resilience in the face of other stressors, such as toxic substances in the environment. The findings pertain as well to the phenomenon of infection tolerance broadly documented in other reservoirs for human disease agents, such as betacoronaviruses and bats (57). Less directly, the results provide for insights about maladaptive responses among humans to microbes, from systemic inflammatory response syndrome (SIRS) to post-infection fatigue syndromes.

Results

LPS experiment and hematology studies

Twenty adult animals each for P. leucopus and M. musculus and equally divided between sexes received by intraperitoneal injection either purified E. coli LPS at a dose of 10 µg per g body mass or saline alone (Table 1). Within 2 h LPS-treated animals of both species displayed piloerection and sickness behavior, i.e. reduced activity, hunched posture, and huddling. By the experiment’s termination at 4 h, 8 of 10 M. musculus treated with LPS had tachypnea, while only one of ten LPS-treated P. leucopus displayed this sign of the sepsis state (p = 0.005).

Within a given species there was little difference between LPS-treated and control animals in values for erythrocytes. But overall the deermice had lower mean (95% confidence interval) hematocrit at 42 (36-48)%, hemoglobin concentration at 13.8 g/dL (12.1-15.5), and mean corpuscular volume for erythrocytes at 49 fL (47-51) than M. musculus with respective values of 56 (51-62)%, 16.1 g/dL (14.6-17.7), and 60 fL (58-62) (p < 0.01). These hematology values for adult CD-1 M. musculus and LL stock P. leucopus in this study were close to what had been reported for these colony populations (18, 94).

In contrast to red blood cells, the mean numbers of white blood cells in the LPS groups in both species were lower than those of control groups (Figure 1). Controls had a mean 4.9 (3.5-6.4) x 103 white cells per µl among M. musculus and 5.8 (4.2-7.4) x 103 white cells per µl among P. leucopus (p = 0.41). For the LPS-treated animals the values were 2.1 (1.5-2.7) x 103 for mice and 3.1 (0.9-5.4) x 103 for deermice (p = 0.39). However, there was difference between species among LPS-treated animals in the proportions of neutrophils and lymphocytes in the white cell population. The ratios of neutrophils to lymphocytes were 0.25 (0.14-0.45) and 0.20 (0.13-0.31) for control M. musculus and P. leucopus, respectively (p = 0.53). But under the LPS condition. the neutrophil-to-lymphocyte ratio was 0.18 (0.11-0.28) for mice and 0.64 (0.42-0.97) for deermice (p = 0.0006). The regression curves for plots of neutrophils and lymphocytes for LPS-treated and control P. leucopus and LPS-treated M. musculus had similar slopes, but the y-intercept was shifted upwards towards a higher ratio of neutrophils to lymphocytes for blood from the LPS group of deermice. Control group mice and deermice and LPS-treated mice had similar percentages (∼5%) of monocytes in their blood; the mean monocyte percentage rose to 10% in LPS treated deermice (p = 0.12). Eosinophil percentages tended to be higher in deermice at a mean 3.4 (2.1-4.7)% than mice at 1.2 (0.5-1.9)% under either condition (p = 0.004).

Total white blood cells, neutrophils, and lymphocytes of Mus musculus (M) and Peromyscus leucopus (P) with or without (control; C) treatment with 10 µg lipopolysaccharide (LPS; L) per g body mass 4 h previous. The box plots of left and center panels show values of individual animals and compiled median, quartiles, and extreme values. The linear regressions of the right panel are color-coded according to the species and treatment designations. The outlier value for a M. musculus control (MM17) was excluded from the linear regression for that group.

In the P. leucopus experiment with a 10-fold lower dose of LPS and a 12 hour duration, the mean (95% confidence interval) white blood cell count (x 103) at termination 3.5 (2.5-4.5) in controls and 7.9 (6.0-9.7) in the LPS-treated (p = 0.01). Even with the higher overall white blood cell count the increase white cells was proportionately higher for neutrophils than for lymphocytes, as was seen in the deermice in the higher dose LPS experiment. The ratio of neutrophils-to-lymphocytes was 0.20 (0.07-0.32) in the controls and 0.38 (0.26-0.50) in the LPS-treated (p = 0.10).

The higher neutrophil to lymphocyte ratio in the deermice exposed to LPS was consistent with the greater neutrophil activation noted by transcriptional analysis (3). But many individual genes that constitute this and related gene ontology (GO) terms had transcription levels in the deermice that far exceeded a three-fold difference in neutrophil counts. For some genes the differences were a hundred or more fold, which suggested that the distinctive LPS transcriptional response profile between species was not attributable solely to neutrophil counts.

Genome-wide expression in blood of deermice and mice

We used the respective transcript sets from the reference genomes for P. leucopus and M. musculus for deep coverage RNA-seq with paired-end ∼150 nt reads (Dryad Tables D1 and D2). Principle component analyses (PCA) of the P. leucopus data and M. musculus data revealed that untreated controls had coherent profiles within each species (Figure 2). With the exception of one mouse, the LPS-treated M. musculus were also in a tight PCA cluster. In contrast, the LPS-treated deermice displayed a diversity of genome-wide transcription profiles and limited clustering.

Principle component analysis of genome-wide RNA-seq data of P. leucopus or M. musculus with or without (blue dot) treatment with LPS 4 h previous. The individual animals listed in Table 1 are indicated on the graphs. The insets indicate the size and color of the symbol for the experimental condition (LPS-treated or control).

For both species the number of genes with higher expression with LPS exposure exceeded those with lower or unchanged expression (Dryad Tables D3 and D4). For P. leucopus and M. musculus the mean fold-changes were 1.32 (1.29-1.35) and 1.30 (1.24-1.36), respectively (p = 0.31). For GO term analysis the absolute fold-change criterion was ≥ 2. Because of the ∼3-fold greater number of transcripts for the M. musculus reference set than the P. leucopus reference set, application of the same false-discovery rate (FDR) threshold for both datasets would favor the labeling of transcripts as DEGs in P. leucopus. Accordingly, the FDR p values were arbitrarily set at <5 x 10-5 for P. leucopus and <3 x 10-3 for M. musculus to provide approximately the same number of DEGs for P. leucopus (1154 DEGs) and M. musculus (1266 DEGs) for the GO term comparison.

Figure 3 shows the GO terms for the top 20 clusters by ascending p-value for up-regulated and down-regulated in P. leucopus and the corresponding categories for M. musculus. The up-regulated gene profile for P. leucopus featured terms associated with “neutrophil degranulation”, “myeloid leukocyte activation”, “leukocyte migration”, and “response to molecule of bacterial origin”. Other sets of up-regulated genes for the deermice were “negative regulation of cytokine production” and “regulation of reactive oxygen species metabolic process”. None of these were among the top 20 up-regulated clusters for M. musculus. Indeed, “leukocyte activation” and “leukocyte migration” were GO terms for down-regulated DEGs in M. musculus. Distinctive GO terms for up-regulated genes distinguishing mice from deermice were “response to virus”, “response to interferon-beta”, “response to interferon-gamma”, “response to protozoan”, and “type II interferon signaling”.

By arbitrary criterion of 100 for the top DEGs by ascending p value for each species, 24 genes were shared between species (Supplementary Materials Table S1). These included up-regulated Bcl3, Ccl3, Cxcl1, Cxcl2, Cxcl3, Cxcl10, Il1rn, and Sod2. Among the 100 mouse DEGs, 20 were constituents of GO terms “response to virus” or “response to interferon-beta” and only 3 were members of GO term sets “response to molecule of bacterial origin” or “response to lipopolysaccharide”. In contrast, among the top 100 deermouse DEGs, there were only 2 associated with the virus or type 1 interferon GO terms, but 12 were associated with either or both of the bacterial molecule GO terms.

We confirmed the sex identification for each sample with sex-specific transcripts of Xist for females and Ddx3y for males (3). For female and male P. leucopus there were 5012 transcripts out of 54,466 in the reference set for which there were TPM values of ≥10 in at least one animal in each of the sexes under either condition (Dryad Table D5). The comparison of females to males by fold-changes between LPS-treated and control animals revealed transcripts that were differentially expressed between sexes under LPS treatment (Table S2 and Figure 4). Some were down-regulated in one sex while unchanged in expression in the opposite sex. Of note in this category were different isoforms or variants of Lilra6 (leukocyte immunoglobulin-like receptor, subfamily A, member 6), one of a family of orphan receptors of myeloid cells (8). The opposite case was exemplified by the Dnajc15 and Hspa8 genes for two chaperones: DnaJ heat shock protein family (Hsp40) member C15 and heat shock protein 8, respectively. These were substantially lower in transcription in the LPS-treated females than in untreated animals, but little changed in LPS-treated males. Coordinates for some other genes, e.g. Saa5 and Cxcl2, fell outside the prediction limits at the extreme end of up-regulation, but their vectors were within 20-25° of each other. While these and other sex-associated differences merit attention for future studies, overall they were not of sufficient number or magnitude in our view to warrant division by sex for the subsequent analyses, which had the aim of identifying differences applicable for both females and males.

Gene Ontology (GO) term clusters associated with up-regulated genes (upper panels) and down-regulated genes (lower panels) of P. leucopus (left panels) and M. musculus (right panels) treated with LPS in comparison with untreated controls of each species. The scale for the x-axes for the panels was determined by the highest -log10 p values in each of the 4 sets. The horizontal bar color, which ranges from white to dark brown through shades of yellow through orange in between, is a schematic representation of the -log10 p values.

Scatter plot with linear regression of pairs of log2-transformed mean fold-changes between LPS-treated and control P. leucopus by male and female sex. The 5012 reference transcripts from the genome reference set are defined in the text and listed in Table S2 and Dryad Table D5. The coefficient of determination (R2), the 95% upper and lower prediction limits for the regression line, and distributions of the values on the x-and y-axes are shown. Selected genes for which their x-y coordinates fall outside the limits of prediction are labeled. Cxcl2, Ibsp, Saa3, Saa5, Sbno2, Serpine1, Slpi, and Steap1 were noted as up-regulated DEGs for the groups with both sexes (Table S1).

Targeted RNA seq analysis

The emerging picture was of P. leucopus generally responding to LPS exposure as if infected with an extracellular bacterial pathogen, including with activated neutrophils. While M. musculus animals of both sexes shared with P. leucopus some features of an antibacterial response, they also displayed type 1 and type 2 interferon type response profiles associated with infections with viruses and intracellular bacteria and parasites.

Going forward, the challenge for a cross-species RNA-seq was commensurability between annotated transcripts of reference sets. Orthologous genes can be identified, but mRNA isoforms and their 5’ and 3’ untranslated regions may not fully correspond. Accordingly, we limited targeted RNA-seq to protein coding sequences of mRNAs for the corresponding sets of P. leucopus and M. musculus sequences.

The 113 mRNA coding sequences, which are listed in Methods, were drawn from the identified DEGs for P. leucopus and M. musculus from the genome-wide RNA-seq. For cross-species normalization we first evaluated three methods: (1) normalization using the ratio of mean total reads for all samples to total reads for a given sample, (2) the ratio of reads mapping to a given target gene (i.e. the numerator) to the reads to the mitochondrial 12S rRNA gene (i.e. the denominator), or (3) when the denominator instead was the myeloid cell marker CD45 (protein tyrosine phosphatase, receptor type C) encoded by the Ptprc gene. In humans, mice, and hamsters, Ptprc is expressed by nucleated hematopoietic cells, and CD45 is commonly used as a white cell marker for flow cytometry (82). The coefficients of determination (R2) between comparison pairs (e.g. normalized total reads vs. Ptprc ratio) within a species were ≥ 0.95 (Supplementary Materials Figure S1 and Table S3). There was also little difference between the choice of Ptprc or 12S rRNA transcripts as denominator with respect to cross-species comparisons of LPS-treated to control fold changes. Given the widespread adoption of CD45 for flow cytometry, we chose Ptprc as denominator and as an adjustment for white cell numbers in the samples. Pearson correlation between log-transformed total white blood cell counts and normalized reads for Ptprc across 40 animals representing both species, sexes, and treatments was 0.40 (p = 0.01).

Figure 5 comprises plots of the log-transformed mean ratios for the 10 P. leucopus controls and 10 M. musculus controls and for the 10 P. leucopus and 10 M. musculus treated with LPS (Table S4). For untreated animals (left panel) there was high correlation and a regression coefficient of ∼1 between the paired data for deermice and mice. The mitochondrial cytochrome oxidase 1 gene (MT-Co1) and S100a9, a subunit of calprotectin, were comparably transcribed. But, there were other coding sequences that stood out for either their greater or lesser transcription in untreated deermice than mice. Two examples of greater expression were Arg1 and MX dynaminin-like GTPase 2 (Mx2), an ISG, while two examples of lesser expression were matrix metalloprotease 8 (Mmp8) and Slpi. There was low to undetectable transcription of Nos2 and interferon-gamma (Ifng) in the blood of controls of both species.

Scatter plots with linear regression of pairs of log-transformed (ln) normalized RNA-seq reads for selected coding sequences for control P. leucopus and M. musculus (left panel) and LPS-treated P. leucopus and M. musculus (right panel). The R2 values and selected genes (each with a different symbol) are indicated in each graph.

For the LPS-treated animals (right panel Figure 5) there was, as expected for this selected set, higher expression of the majority genes and greater heterogeneity among P. leucopus and M. musculus animals in their responses for represented genes. In contrast to the findings with controls, Ifng and Nos2 had higher transcription in treated mice. In deermice the magnitude of difference in the transcription between controls and LPS-treated was less. A comparatively restrained transcriptional response in deermice was also noted for Mx2. On the other hand, there were greater fold-change from baseline in P. leucopus than in M. musculus for interleukin-1 beta (Il1b), Mmp8, Slpi, and S100a9.

Table 2 lists all the selected targets with the means and confidence intervals for the normalized values for controls and LPS-treated M. musculus and controls and LPS-treated P. leucopus (Table S4). Box plots for a selected 54 of these targets organized by functional characteristics are provided in Figure S2. The fold-changes within each species and between treatments across species are given. The final column is the ratio of the fold-change between LPS to control in P. leucopus to the corresponding value for M. musculus. This along with the derived heat-map of these ratios, presented in the second column, indicates the genes for which there was little difference between species in their responses to LPS--either up or down--as well as those that were comparatively greater or lesser in one species or the other. Several of these genes are considered in other specific contexts below. Of note here are the places of Nos2 and Ifng at the bottom of the table, and Il1b near the top at position 20.

“Alternatively-activated” macrophages and “nonclassical” monocytes in P. leucopus

While we could not type single cells using protein markers, we could assess relative transcription of established indicators of different white cell subpopulations in whole blood. The present study, which incorporated outbred M. musculus instead of an inbred strain, confirmed the previous finding of differences in Nos2, the gene for inducible nitric oxide synthase and Arg1, the gene for arginase 1, expression between M. musculus and P. leucopus (Figure 5; Table 2). Results similar to the RNA-seq findings were obtained with specific RT-qPCR assays for Nos2 and Arg1 transcripts for P. musculus and M. musculus (Table 3).

Low transcription of Nos2 in both in controls and LPS-treated P. leucopus and an increase in Arg1 with LPS was also observed in another experiment for present study where the dose of LPS was 1 µg/g body mass instead of 10 µg/g and the interval between injection and assessment was 12 h instead of 4 h (Table 4).

In addition to the differences in Nos2 and Arg1 expression for typing macrophage and monocyte subpopulations, there are also the relative expressions of three other pairs of genes: (1) IL12 and IL10, where a lower IL12/IL10 ratio is more characteristic of alternatively activated or M2 type (68, 90); (2) the proto-oncogene kinases Akt1 and Akt2, where the associations are Akt1 with M2-type and Akt2 with M1-type macrophages (1, 91); and (3) CD14 and CD16 (low affinity immunoglobulin gamma Fc region receptor III; Fcgr3), where low expression of CD14 and high expression of CD16/Fcgr3 is associated with “non-classical” monocytes (70). There is evidence that nonclassical monocytes can change to M2-type macrophages (42).

These four relationships, which are presented as log-transformed ratios of Nos2/Arg1, IL12/IL10, Akt1/Akt2, and CD14/CD16, are shown in Figure 6. We confirmed the difference between P. leucopus and M. musculus in the ratios of Nos2/Arg1 and IL12/IL10 (3) with outbred mice and normalization for white cells. In both species the Akt1/Akt2 ratio declined in LPS-treated animals, but for P. leucopus the ratio remained > 1.0 even among LPS-treated animals, while in the blood of M. musculus the ratio was < 1.0 at baseline and declined further in the LPS-treated animals.

Box plots of log-transformed ratios of four pairs of gene transcripts from targeted RNA-seq analysis of blood of P. leucopus (P) or M. musculus (M) with (L) or without (C) treatment with LPS. The values are from Table S4. Upper left, Nos2/Arg1. Upper right, IL12/IL10. Lower left, Akt1/Akt2. Lower right, Cd14/Cd16.

An ortholog of Ly6C (13), a protein used for typing mouse monocytes and other white cells, has not been identified in Peromyscus or other Cricetidae family members. Therefore, for this study the comparison with Cd14 is with Cd16 or Fcgr3, which deermice and other cricetines do have. In mice the Cd14/Cd16 ratio increased from baseline in the LPS group. In the deermice the ratio in control animals was midway between the two groups of mice but there was a marked decrease in the LPS-treated deermice (Figure 6). This was not associated with a fall in the absolute numbers or percentages of monocytes in the blood of these animals.

Taken together, the Nos2-Arg1, Il12-Il10, Akt1-Akt2, and CD14-CD16 relationships document a disposition toward alternatively-activated macrophages and nonclassical monocytes in P. leucopus both before and after exposure to LPS. This contrasts with profiles consistent with a predominance of classically-activated macrophages and classical monocytes in mice.

Interferon-gamma and interleukin-1 beta dichotomy between deermice and murids

For mice the Ifng transcript was one of the top ranked DEGs by both fold-change and adjusted p value by genome-wide RNA-seq (Table 2). In contrast, for P. leucopus Ifng was far down the list, and the comparably ranked DEG instead was Il1b. This inversion of relationships between two pro-inflammatory cytokines was confirmed by analysis of the individual animals of both species (Figure 7). There was little or no detectable transcription of Ifng in the blood of deermice in which Il1b expression was high. There was also scant to no transcription of Ifng in the blood of P. leucopus 12 h after injection of LPS (Table 4).

Transcripts of interferon-gamma and interleukin-1 beta by targeted RNA-seq of the blood of P. leucopus (P) or M. musculus (M) with (L) or without (C) treatment with LPS. The top panels are box plots of the individual values. The lower left panel is a scatter plot of interleukin-1 ý on interferon-ψ values. The lower right panel is a Discriminant Analysis of these pairs of values where Factor 1 corresponds to interferon-gamma, and Factor 2 corresponds to interleukin-1 beta. Values for analysis are in Table S4.

The up-regulation of Ifng within 4 hours of exposure to LPS was not limited to the species M. musculus. In an experiment with the rat R. norvegicus, we used two different LPS doses (5 µg/g and 20 µg/g), but the same 4 h endpoint and whole blood as the sample. Both groups of LPS-treated rats had lowered total white blood cells and, like the mice, lower neutrophil-to-lymphocyte ratios compared to controls (Table 5). There were also elevations of interferon-gamma, interleukin-6 and interleukin-10 proteins from undetectable levels in the blood of the treated rats (Dryad Table D5). The values for rats receiving 5 µg/g or 20 µg/g doses were similar (Figure S3), so these groups were combined. By targeted RNA-seq there were 24x fold-changes between the LPS-treated rats and control rats for Ifng and Nos2 but only ∼3x fold-change for Il1b (Table 5).

We asked why the Infg response observed in CD-1 mice and rats here was not as pronounced in BALB/c mice (3). Accordingly, we used the RNA-seq reads from the prior study in combination with the reads of the present study and carried out targeted RNA-seq (Figure S4). The BALB/c inbred mice had, like the CD-1 mice, modest elevations of Il1b transcription. Ifng expression was also elevated in the BALB/c animals but not to the degree noted in CD-1 mice or rats. One explanation is an inherent difference of BALB/c mice from other strains in their lower interferon-gamma response to LPS (47, 84).

Interferon-gamma and inducible nitric oxide synthase

Interferon-gamma is a determinant of Nos2 expression (55, 81). So, the scant transcription of Ifng in P. leucopus conceivably accounted for the low expression of Nos2 in that species. The analysis shown in upper left panel of Figure 8 shows a tight correlation between the levels of transcription of Ifng and Nos2 for both species and both experimental conditions (Figure S5). A significant correlation was also observed for the combined set of animals between the ratios of Nos2 to Arg1 and Ifng to Il1b (upper right panel), an indication of co-variation between Ifng expression and macrophage polarization.

Normalized transcripts of Nos2, Ifng, and Cd69 in targeted RNA-seq analysis of blood of P. leucopus (P) or M. musculus (M) with (L) or without (C) treatment with LPS. Upper left: scatter plot of individual values for Nos2 on Ifng with linear regression curve and coefficient of determination (R2). Upper right: natural logarithm (ln) of ratios of Nos2/Argi1 on Ifng/IL1b with regression curve and R2. Lower left: Box plots of individual values of normalized transcripts of Cd69. Lower right: Scatter plot of Ifng on Cd69 with separate regression curves and R2 values for M. musculus and P. leucopus. Values for analysis are in Table S4. Box plots for Nos2 and Arg1 are provided in Figure S5, and box plots for Ifng and Il1b are provided in Figure 7.

The plausible sources of Ifng mRNA in whole blood are T-cells, Natural Killer cells, and Type 1 Innate Lymphoid Cells (75). A DEG for M. musculus by both genome-wide and targeted RNA-seq (Table 2; Table S1) was Cd69, a C-type lectin protein and an early activation antigen for these cells(38). In P. leucopus transcription Cd69 occurred in the blood of control P. leucopus, but it was the same or only marginally different for the LPS-treated animals (lower left panel). In contrast, in M. musculus the baseline transcription of Cd69 was below that of P. leucopus, but in the LPS-treated mice it was many fold higher. In mice transcripts for Cd69 correlated tightly with Ifng transcription, while in the deermice there was little correlation between Cd69 and Ifng expression at those low levels (lower right panel).

The findings are consistent with CD69-positive cells being a source of Ifng in mice. Cd69 transcription was comparatively higher in control deermice than in control mice, so we presume that deermice have CD69-positive cells at baseline. One explanation then for the comparatively few Ifng transcripts in the deermice after LPS is a diminished responsiveness of these cells. Tlr4 expression increased ∼3 -fold more in P. leucopus than in M. musculus after LPS (Table 2), but the magnitude of the decline in expression of Cd14 in deermice than mice was even greater. CD14 is required for LPS-stimulated signaling through surface TLR4 (58), and, as such, its decreased availability for this signaling pathway is a possible explanation for the moderated response to LPS in P. leucopus.

Interferon-stimulated genes and RIG-I-like receptors

As noted, GO terms differentiating mice from deermice included “response to interferon-beta” and “response to virus” (Figure 3). There was also the example of Mx2, an ISG with anti-viral activity on its own, that showed a greater fold-change from baseline in mice than in deermice (Figure 5). Five other ISGs--guanylate binding protein 4 (Gbp4), interferon-induced protein with tetratricopeptide repeat (Ifit1), interferon regulatory factor 7 (Irf7), ubiquitin-type modifier ISG15 (Isg15), and 2’-5’ oligoadenylate synthase 1A (Oas1a)—had higher transcription in all the LPS-treated animals. But the magnitude of fold change was less in the deermice, ranging from 6-25% of what it was in the LPS group of mice (Table 2; Figure S6).

The up-regulation of these ISGs was evidence of an interferon effect, but transcripts for interferon-1 beta (Ifnb) or -alpha (Ifna) themselves were scarcely detectable in deermice or mice in the blood under either condition. We then considered pattern recognition receptors (PRR) that might be part of a signaling pathway leading to ISG expression. Among the DEGs from the genome-wide analyses were four cytoplasmic PRRs: (1) Rigi (formerly Ddx58), which encodes the RNA helicase retinoic acid-inducible I (RIG-I); (2) Ifih1, which encodes interferon induced with helicase C domain 1, also known as MDA5 and a RIG-I-like receptor; (3) Dhx58, also known LGP2 and another RIG-I-like receptor; and (4) cGAS (cyclic GMP-AMP synthase), which is part of the cGAS-STING sensing pathway (Table S4).

All four cytoplasmic PRRs were upregulated in the blood of LPS-treated mice and deermice (Table 2). But, again, for each of them the magnitude of fold change was less by 50-90% in treated P. leucopus than in M. musculus. The coefficients of determination for the 6 ISGs and the 4 PRRs are provided in Figure 9. For most of the pairs there was evidence of covariation across all 40 animals. When the correlation was low across all the data, e.g. between the ISG Mx2 and the PRR Rigi or the ISGs Mx2 and Gbp4, it was high within a species.

Co-variation between transcripts for selected PRRs and ISGs in the blood of P. leucopus (P) or M. musculus (M) with (L) or without (C) LPS treatment. Top panel: matrix of coefficients of determination (R2) for combined P. leucopus and M. musculus data. PRRs are indicated by yellow fill and ISGs by blue fill on horizontal and vertical axes. Shades of green of the matrix cells correspond to R2 values, where cells with values less than 0.30 have white fill and those of 0.90-1.00 have deepest green fill. Bottom panels: scatter plots of log-transformed normalized Mx2 transcripts on Rigi (left), Ifih1 (center), and Gbp4 (right). The linear regression curves are for each species. For the right-lower graph the result from the General Linear Model (GLM) estimate is also given. Values for analysis are in Table S4; box plots for Gbp4, Irf7, Isg15, Mx2, and Oas1 are provided in Figure S6.

These findings were evidence that pathways in P. leucopus for PRR signaling and ISG expression functioned similarly to those in M. musculus but differed under these experimental conditions in magnitude of the changes, being more moderate in the deermice.

Endogenous retroviruses in deermice, mice, and rats after LPS exposure

The six ISGs are nonexclusive consequences of activity of type 1 interferons. What we could document was the association of transcription of the gene for the cytoplasmic PPRs, including RIG-I, and the ISGs in both species, as well as the distinction between deermice in the magnitude of the responses of both PRRs and ISGs. These findings led us to ask could be a pathogen-associated molecular pattern (PAMP) for signaling pathways leading to expression of type 1 interferons.

One of these is endogenous retroviruses (ERV). The activity of these diverse, abundant, and pervasive elements have been recognized as one of the drivers of innate immune responses to a microbe (41, 51, 76). Our attention was drawn to ERVs by finding in the genome-wide RNA-seq of LPS-treated and control rats. Two of the three highest scoring DEGs by FDR p value and fold-change were a gag-pol polyprotein of a leukemia virus with 131x fold-change from controls and a mouse leukmia virus (MLV) envelope (Env) protein with 62x fold-change (Dryad Table D5).

We returned to the mouse and deermouse data. There were four MLV or other ERV Env proteins among the 1266 genome-wide RNA-seq DEGs for M. musculus. But, there was no ERV Env protein identified as such among the 1154 DEGs identified for P. leucopus (Table S1). One possible explanation for the difference was an incomplete annotation of the P. leucopus genome. We took three approaches to rectify this. The first was to examine the DEGs for P. leucopus that encoded a polypeptide ≥ 200 amino acids and was annotated for the genome as “uncharacterized”. A search with these candidates of both the virus and rodent proteins databases identified two that were homologous with gag-pol polyproteins of ERVs, mainly leukemia viruses, of mammals.

For a second approach we carried out a de novo transcript assembly of mRNA reads from blood of LPS-treated and control P. leucopus and used the resultant contigs as the reference set for RNA-seq analysis. This identified two contigs that were measurably transcribed in the blood, differentially expressed between conditions, and were homologous to ERV sequences. One was revealed to be an Env protein that was identical to a P. leucopus protein annotated as “MLV-related proviral Env protein” (XP_037065362). The second was a gag-pol protein. The latter was near-identical to the gag-pol protein identified by the first approach.

The third approach was to scan the P. leucopus genome for nonredundant sequences, defined as <95% identity, that were homologous with ERV gag-pol sequences, which are not typically annotated because of masking for repetitive sequences. This analysis yielded 615 unique sequences. These were used in turn as a reference set for RNA-seq. There were 4 sequences that met the criterion of FDR p value <0.01. Three were transcribed at 5-to 40-fold higher levels in LPS-treated deermice than in controls. But all three, as well as the fourth, a down-regulated DEG, were ERV relics with truncations, frame shifts, and in-frame stop codons. These were assessed as non-coding RNAs and not further pursued in this study.

To represent P. leucopus in a targeted RNA-seq comparison with mice and rats we settled on the Env protein and gag-pol coding sequences identified present in the blood mRNA and as DEGs. Representing M. musculus were highest ranked MLV Env and gag-pol protein sequences among the DEGs. For rats we chose Env and gag-pol proteins that were second and third ranked DEGs identified in the genome-wide RNA-seq. Because of length differences for the coding sequences, the unit used for cross-species analysis was reads per kilobase before normalization for Ptprc transcription.

The left panel of Figure 10 shows the striking transcriptional fold-change in LPS-treated rats of both Env and gag-pol sequences over controls. Of lesser magnitude but no less significant was the fold-change observed M. musculus for both Env and gag-pol sequences. In both mice and rats Env and gag-pol read values were highly correlated across conditions. In contrast, in P. leucopus the magnitudes of fold-change upwards for gag-pol was less than in mice or rats, and transcription of the Env protein sequence was actually lower in LPS-treated animals than in controls. While there was a tight association between Env protein and the PRR Rigi transcription in the M. musculus, this was not observed in P. leucopus. Rigi transcription was moderately higher at the time that Env protein’s transcription was lower in the LPS group.

Scatter plots of endogenous retrovirus (ERV) Env and Gag-pol protein gene transcription (left) and association of ERV Env with Rigi transcription (right) in the blood of P. leucopus (Pero; P), M. musculus (Mus; M), or R. norvegicus (Rattus) with (L) or without (control; C) treatment with LPS. In right panel the linear regression curve and coefficients of determination (R2) for P. leucopus and M. musculus are shown. Values for analysis are in Table S4; box plots for ERV Env and ERV Gag-pol are provided in Figure S7.

Borrelia hermsii infection of P. leucopus

The phenomena reported so far were consequences of exposures to a particular PAMP--bacterial lipopolysaccharide with its hallmark lipid A moiety--recognized by a particular PRR, TLR4. While the focus was primarily on events downstream from that initial signaling, we asked in a concluding study whether the profile observed in P. leucopus applied in circumstances when the PAMP or PAMPs did not include LPS. This question is germane, given P. leucopus’ role as a natural host for B. burgdorferi. This organism and other members of the spirochete family Borreliaceae do not have LPS (5, 88), but they have abundant lipoproteins, which are agonists for TLR2 in a heterodimer with TLR1 (80). B. burgdorferi is transiently blood-borne at low densities in P. leucopus, but in its life cycle B. burgdorferi is mainly tissue-associated in vertebrate hosts (6). We previously observed that the blood of B. burgdorferi-infected P. leucopus manifested few DEGs in comparison to skin (52). More comparable to the LPS experimental model is infection of P. leucopus with a relapsing fever Borrelia species, which commonly achieve high densities in the blood. P. leucopus is a reservoir for Borrelia miyamotoi, which causes hard tick-borne relapsing fever (6), and the related P. maniculatus is a natural host for the soft tick-borne relapsing fever agent B. hermsii (43).

Accordingly, we used blood RNA-seq reads, which were taken from a prior study of B. hermsii infection of P. leucopus (3), for targeted analysis with the same reference set employed for the LPS analyses (Table 6 and Table S5). The blood samples were taken from infected and uninfected animals on day 5, when bacteremia was at its peak, as documented by microscopy of the blood, qPCR of the spleen, and transcripts of a B. hermsii plasmid in the RNA extracts of the blood. As expected for B. hermsii infection (24), the spleen was enlarged in infected animals.

Similarities in the profiles for the LPS-treated and B. hermsii-infected deermice were as follows: (1) low levels of transcription of Nos2 and Ifng that contrasted with the high levels for Arg1 and Il1b in the same animals, (2) maintenance of Akt1/Akt2 > 1.0 under both conditions, (3) reduction of the Cd14/Cd16 ratio, (4) decreased transcription of Cd69, and (5) stable, low transcription of ERV Env and gag-pol loci with only marginal increases in transcription of ISGs and RIG-I-like receptors. Other equivalences under the two experimental conditions included increases in expression of genes for superoxide dismutase 2, low-affinity Fc gamma receptors, and secretory leukocyte protease inhibitor. Thus, the responses that distinguish deermice are not confined to the singular case of LPS as the elicitor.

Discussion

Study limitations

The approach was forward and unbiased, looking for differences between species broadly across their transcriptomes. The findings lead to hypotheses, but reverse genetics in service of that testing was not applied here. In selective cases we could point to supporting evidence in the literature on M. musculus and the phenotypes of relevant gene knockouts, but there are no such resources for Peromyscus as yet. The resource constraint also applies to the availability of antibodies for use with Peromyscus for immunoassays for specific proteins, e.g. interferon-gamma, in serum, or for cell markers, e.g. CD69, for flow cytometry of white blood cells.

While a strength of the study was use of an outbred population of M. musculus to approximate the genetic diversity of the P. leucopus in the study, this meant that some genes of potential relevance might have gone undetected, i.e. from type II error. The variances for a sample of genetically diverse outbred animals, like the LL stock of P. leucopus (52, 53), would be expected to be greater than for the same sized sample of inbred animals. For some traits, especially ones that are complex or under balancing selection, even sample sizes of 10 in each group may not have provided sufficient power for discrimination between deermice and mice. For the same reason differences between sexes of a species in their responses might have been undetected. The interpretations applied to mixed-sex groups of deermice and mice. Expression strongly associated with female or male sex could have yielded an average fold change for the whole group that fell below the screen’s threshold.

The parameters for the experiment of LPS dose, the route, and duration of experiment each might have had different values under another design. Those particular choices were based on past studies of deermice and mice (3, 49). In another experiment we found that with doses twice or half those given the deermice the responses by rats to the different doses were indistinguishable by hematology, cytokine assays, and RNA-seq. Thus, there seems to be some latitude in the dose and still achieving replication. We obtained similar results for P. leucopus when we looked at a replicate of the experiment with the same conditions (3), when the dose was lower and duration lengthened to 12 h (this study). The analysis here of the B. hermsii infection experiment also indicated that the phenomenon observed in P. leucopus was not limited to a TLR4 agonist.

While the rodents in these experiments were housed in the same facility and ate the same diet, we cannot exclude inherent differences in gastrointestinal microbiota between species and individual outbred animals as co-variables for the experimental outcomes. We reported differences between the LL stock P. leucopus and BALB/c M. musculus of the same age and diet in their microbiomes by metagenomic analysis and microbiologic means (61). This included a commensal Tritrichomonas sp. in P. leucopus but not in the M. musculus in the study. The presence of these protozoa affects innate and adaptive immune responses in the gastrointestinal tract (19, 29), but it is not clear whether there are systemic consequences of colonization by this flagellate.

LPS, ERVs, and interferons

The results confirm previous reports of heightened transcription of ERV sequences in mice or mouse cells after exposure to LPS (35, 44, 87). Here we add the example of the rat. The LPS was administered in solution and not by means of membrane vesicles. The sensing PRR presumably was surface-displayed, membrane-anchored TLR4 (58). It follows that a second, indirect of LPS on the mouse is through its provocation of increased ERV transcription intracellularly. ERV-origin RNA, cDNA and/or protein would then be recognized by a cytoplasmic PRR. RIG-I was one associated with ERV transcription in this study. Kong et al. reported that LPS stimulated expression of Rigi in a mouse macrophages but did not investigate ERVs for an intermediary function in this phenomenon (46). As was demonstrated for LINE type retrotransposons in human fibroblasts, intracellular PRR signaling can trigger a type 1 interferon response (25). The combination of these two signaling events, i.e. one through surface TLR4 by LPS itself and another through intracellular PPR(s) by to-be-defined ERV products, manifested in mice and rats as a response profile that had features of both a response to a virus with type 1 interferon and ISGs and a response to a bacterial PAMP like LPS with acute phase reactants such as calprotectin and serum amyloid.

This or a similar phenomenon has been observed under other circumstances. In humans there was heightened transcription of retrotransposons in patients with septic shock (63), as well as in peripheral blood mononuclear cells from human subjects experimentally injected with LPS (74). Bacteria like Staphylococcus epidermidis that express TLR2 agonists, such as lipoteichoic acid, promoted expression of ERVs, which in turn modulated host immune responses (51). A synthetic analog of a B. burgdorferi lipoprotein activated human monocytic cells and promoted replication of the latent HIV virus in cells that were persistently infected (72).

P. leucopus does not fit well with this model. Instead of the prominent interferon-gamma response observed in mice and rats, there were prominent responses of interleukin-1 beta and genes associated with neutrophil activation. Instead of the much heightened expression of ISGs, like Mx2 and Isg15, in mice treated with LPS, the deermice under the same condition had a more subdued ISG transcription profile. Instead of increased expression of ERV Env protein sequences in blood of mice and rats treated with LPS, there was decreased transcription of the homologous MLV Env gene in like-treated P. leucopus.

This suppression in the deermice may be attributable to defensive adaptations of Peromyscus to repeated invasions of endogenous retroviruses, as Gozashti et al. has proposed for P. maniculatus (32). This includes expanding the repertoire of silencing mechanisms, such as Kruppel-associated box (KRAB) domain-containing zinc finger proteins (97). Like P. maniculatus, P. leucopus has an abundance of Long Terminal Repeat retrotransposons, several named for their endogenous retrovirus heritages (52). Our initial analysis of the P. leucopus genome reported a depletion of KRAB domains compared to Muridae (52). But a subsequent annotation round identified several genes for KRAB domain zinc finger proteins in P. leucopus, including Zfp809 (XP_006982432), which initiates ERV silencing (95), and Zfp997 (XP_037067826), which suppresses ERV expression (89). Another possible adaptation in P. leucopus is the higher baseline expression of some ISGs as noted here (Figure 9 and Figure S6).

Reducing differences between P. leucopus and murids M. musculus and R. norvegicus to a single attribute, such as the documented inactivation of the Fcgr1 gene in P. leucopus (7), may be fruitless. But the feature that may best distinguish the deermouse from the mouse and rat is its predominantly anti-inflammatory quality. This characteristic likely has a complex, polygenic basis, with environmental (including microbiota) and epigenetic influences. An individual’s placement is on a spectrum or, more likely, a landscape rather than in one or another binary or Mendelian category.

One argument against a purely anti-inflammatory characterization is the greater neutrophil numbers and activity in P. leucopus compared to M. musculus in the LPS experiment. The neutrophil activation, migration, and phagocytosis would be appropriate early defenses against a pyogenic pathogen. But if not contained, they bring local and systemic risks for the host. This damage would not likely be from nitric oxide and reactive nitrogen species, given the minimal Nos2 transcription. But deermice showed heightened expression of proteases, such as Mmp8, enzymes for reactive oxygen species, such as NADPH oxidase 1 (Nox1), and facilitators of neutrophil extracellular traps, such as PAD4 (Padi4) (Table 2). We had previously identified possible mitigators, such as the protease inhibitor Slpi and superoxide dismutase 2 (3). These findings were replicated here. The topic of neutrophil activation and these and other possible counters is considered in more detail elsewhere.

An anti-inflammatory disposition but at what cost?

An assignment of infection tolerance to a host and pathogen pairing assumes sufficient immunity against the microbe to keep it in check if elimination fails. P. leucopus and P. maniculatus, are in this sense “immunocompetent” with respect to the microbes they host and with which they may share a long history (40). Yet, has this balance of resistance and tolerance for certain host-associated microbes been achieved in a trade-off that entails vulnerabilities to other types of agents?

The selection of LPS as the experimental model was meant to cover this contingency, at least for the common denominator of acute inflammation many types of infections elicit. But LPS studies revealed potential weaknesses that some pathogens might exploit. One of these is the low expression of inducible nitric oxide. Although Nos2 gene knockouts in M. musculus had lower LPS-induced mortality than their wild-type counterparts, the mutants were more susceptible to the protozoan Leishmania major and the facultative intracellular bacterium Listeria monocytogenes (56, 93). While there are no known studies of either of these pathogens in P. leucopus, the related species P. yucatanicus is the main reservoir for Leishmania mexicana in Mexico (17). Compared with M. musculus, which suffer a high fatality rate from experimental infections with L. mexicana, P. yucatanicus infections are commonly asymptomatic (54).

Given the restrained interferon and ISG response shown by P. leucopus, another plausible vulnerability would be viral infections. But other studies suggest that neither RNA nor DNA viruses pose an inordinately high risk for Peromyscus. Both tolerance of and resistance to the tickborne encephalitis flavivirus Powassan virus by P. leucopus were demonstrated in an experimental model in which mice, by contrast, gwere severely affected (62). P. maniculatus has been successfully infected with the SARS-CoV-2 virus by the respiratory route, but the infected animals displayed only mild pathology, manifested little if any disability, and recovered within a few days (30, 33). Among natural populations and in the laboratory P. maniculatus is noted for its tolerance of hantavirus, which commonly is fatal for infected humans (14, 20). P. maniculatus was permissive of infection with monkeypox virus, but the infection was mild and transient (26).

A distinguishing P. leucopus characteristic, which was not expressly examined here, is its aforementioned 2-3 fold greater life span than that of M. musculus. While deermice may not be in the same longevity league as the naked mole-rat (Heterocephalus glaber), which can live for over 30 years (73), some features of naked mole-rat immunology are intriguingly similar to what we have observed for P. leucopus. These include macrophages and blood myeloid cells with low to absent transcription of Nos2 or production of nitric oxide in response to LPS, even in the presence of added inteferon-gamma (31). Like P. leucopus and in distinction to M. musculus, naked mole-rats showed an increase in the proportion of neutrophils in the blood 4 hours after intraperitoneal injection of LPS (39). In another comparative study, the hematopoietic stem and progenitor cells of these rodents had a lower type 1 interferon response than mice to a TLR3 agonist (28).

In summary, if there is a vulnerability that Peromyscus accepts in return for relief from inflammation (and perhaps a longer life), it has not been identified yet. However, potential threats and stressors are many, and the number assessed either in the field or laboratory has been limited to date.

Implications for Lyme disease and other zoonoses

Our studies of P. leucopus began with a natural population and documented a >80% prevalence of infection and high incidence of re-infections by B. burgdorferi in the area’s white-footed deermouse, the most abundant mammal there (16). This was a Lyme disease endemic area (27), where residents frequently presented for medical care for a variety of clinical manifestations, from mild to serious, of B. burgdorferi infection (85). Subclinical infections in humans occur, but most of those who become infected have a definable illness (86). The localized or systemic presence of the microbe is a necessary condition for Lyme disease, but the majority of the symptoms and signs are attributable to inflammation elicited by the organism’s presence and not from virulence properties per se or the hijacking of host cells (21). Since humans are transmission dead-ends for B. burgdorferi and many other zoonotic agents in their life cycles, it is not surprising that human infections are generally more debilitating if not fatal than what adapted natural hosts experience.

It is in the space between the asymptomatic natural host and symptomatic inadvertent host where there may be insights with basic and translational application. With this goal, we consider the ways the results inform studies of the pathogenesis of Lyme disease, where “disease” includes lingering disorders akin to “long Covid” (71), and where “pathogenesis” includes both microbial and host contributions. Plausibly-germane deermouse-mouse differences identified in our studies are summarized in Figure 11. Two are highlighted here.

Summary of distinguishing features of transcriptional responses in the blood between P. leucopus and M. musculus 4 h after treatment with LPS. There is semi-quantitative representation of relative transcription of selected coding sequences or ratios of transcription for selected pairs of genes in the blood.

The first is macrophage polarization (68). By the criteria summarized above, the response to LPS by P. leucopus is consistent with the alternatively-activated or M2 type, rather than the expected classical or M1 type. But it was not only LPS-treated deermice that had this attribute, the blood of untreated animals also displayed M2 type polarization features. This included a comparatively high Arg1 expression level and a Akt1/Akt2 ratio of more than 1 at baseline. This suggests that studies of other mammals, including humans, need not administer LPS or other TLR agonist to assess disposition toward M1 or M2-type polarization. This reading could serve as a prognostic indicator of the inflammatory response to infection with B. burgdorferi or other pathogen and the long-term outcome.

The second difference we highlight is the activation of ERV transcription that was prominent in the LPS-treated mice and rats but not in similarly-treated deermice. A paradoxical enlistment of antiviral defenses, including type 1 and type 2 interferons, for an infection with an extracellular bacterium, like B. burgdorferi, may bring about more harm than benefit, especially if the resultant inflammation persists after antibiotic therapy. There are various ways to assess ERV activation in the blood, including assays for RNA, protein, and reverse transcriptase activity. A xenotropic MLV-related retrovirus has been discounted as a cause of chronic fatigue syndrome (60). However, production of whole virions need not occur for there to be PRR signaling in response to cytoplasmic Env protein, single stranded RNA, or cDNA (78).

Methods

Animals

The study was carried out in accordance with the Guide for the Care and Use of Laboratory Animals: Eighth Edition of the National Academy of Sciences. The protocols AUP-18-020 and AUP-21-007 were approved by the Institutional Animal Care and Use Committee of the University of California Irvine.

Peromyscus leucopus, here also referred to as “deermice”, were of the outbred LL stock, which originated with 38 animals captured near Linville, NC and thereafter comprised a closed colony without sib-sib matings at the Peromyscus Genetic Stock Center at the University of South Carolina (45). The LL stock animals for this study were bred and raised at the vivarium of University of California Irvine, an AAALAC approved facility. Outbred Mus musculus breed CD-1, specifically Crl:CD1(ICR) IGS, and here also referred to as “mice”, were obtained from Charles River. Fischer F344 strain inbred Rattus norvegicus, specifically F344/NHsd, and here also referred to as “rats”, were obtained from Charles River.

For the combined P. leucopus-M. musculus experiment the 20 P. leucopus were of a mean (95% confidence interval) 158 (156-159) days of age and had a mean 21 (19-22) g body mass. The 20 M. musculus were all 149 days of age and had a mean body mass of 47 (43-50) g. The ratio of average male to average female body mass was 1.04 for P. leucopus and 1.03 for M. musculus. The 6 female P. leucopus for the 12 h duration experiment were of a mean 401 (266-535) days of age and mean body mass of 20 (17-23) g. The 16 adult 10-12 week old female R. norvegicus had a mean 139 (137-141) g body mass. The 7 male P. leucopus for the infection study were of a mean 107 (80-134) days and mean body mass of 21 (18-24) g.

Animals were housed in Techniplast-ventilated cages in vivarium rooms with a 16 h-8 h light-dark cycle, an ambient temperature of 22 °C, and on ad libitum water and a diet of 2020X Teklad global soy protein-free extruded rodent chow with 6% fat content (Envigo, Placentia, CA).

For injections the rodents were anesthetized with inhaled isoflurane. The rodents were euthanized by carbon dioxide overdose and intracardiac exsanguination at the termination of the experiment. No animals died or became moribund before the 4 hour or 12 h termination time points in the LPS experiments or before the 5 d termination point of infection study.

LPS experiments

For the P. leucopus and M. musculus combined experiment, sample sizes replicated the specifications of the previous study, in which there were 20 P. leucopus and 20 M. musculus, equally divided between females and males and equally allotted between conditions (3). The treatments were administered in the morning of a single day. At 15 min intervals and alternating between species, sex, and treatments, animals were intraperitoneally (ip) injected 50 µl volumes of either ion-exchange chromatography-purified Escherichia coli O111:B4 LPS (Sigma-Aldrich L3024) in a dose of 10 µg per g body mass or the diluent alone: sterile-filtered, endotoxin-tested, 0.9% sodium chloride (Sigma-Aldrich). The animals were visually monitored in separate cages continuously for the duration of the experiment. We recorded whether there was reduced activity by criterion of huddling with little or movement for > 5 min, ruffled fur or piloerection), or rapid respiration rate or tachypnea. At 4.0 h time after injection animals were euthanized as described above, and sterile dissection was carried out immediately.

Lower dose and longer duration experiment. In an experiment with 6 P. leucopus, the animals were administered the same single dose of LPS but at 1.0 µg/g and the same control solution. The animals were euthanized 12 h after the injection the following day.

Rat LPS experiment. The same experimental design was used for the rats as for the combined deermice-mice experiment, with the exception that the formulation of the E. coli O111:B4 LPS was “cell culture grade” (Sigma-Aldrich L4391), and the groups were sterile saline alone (n = 5), 5 µg LPS per g body mass (n = 6), or 20 µg LPS per g (n = 5).

Experimental infection

The infection of a group of P. leucopus LL stock with the relapsing fever agent B. hermsii and the processing of blood and tissues for RNA extraction 5 days into the infection were described previously (3). In brief, animals were infected intraperitoneally on day 0 with either phosphate-buffered saline alone or 103 cells of B. hermsii MTW, a strain that is infectious for Peromyscus species (43). Bacteremia was confirmed by microscopy on day 4, and the animals were euthanized on day 5. For that prior study the RNA-seq analysis was limited to the genome-wide transcript reference set. For the present study we used the original fastq format files for targeted RNA-seq as described below.

Hematology and plasma analyte assays

For the combined P. leucopus-M. musculus experiment automated complete blood counts with differentials were performed at Antech Diagnostics, Fountain Valley, CA on a Siemens ADVIA 2120i with Autoslide hematology instrument with manual review of blood smears by a veterinary pathologist. For the 12 h duration P. leucopus experiment hematologic parameters were analyzed on an ABCVet Hemalyzer automated cell counter instrument at U.C. Irvine. For the rat experiment complete blood counts with differentials were performed at the Comparative Pathology Laboratory of the University of California Davis. Multiplex bead-based cytokine protein assay of the plasma of the rats was performed at Charles River Laboratories using selected options of the Millipore MILLIPLEX MAP rat cytokine/chemokine panel.

RNA extraction of blood

After the chest cavity was exposed, cardiac puncture was performed through a 25 gauge needle into a sterile 1 ml polypropylene syringe. After the needle was removed, the blood was expelled into Becton-Dickinson K2E Microtainer Tubes, which contained potassium EDTA. Anticoagulated blood was split into a sample that was placed on ice for same-day delivery to the veterinary hematology laboratory and a sample intended for RNA extraction which was transferred to an Invitrogen RiboPure tube with DNA/RNA Later and this suspension was stored at −20 °C. RNA was isolated using the Invitrogen Mouse RiboPure-Blood RNA Isolation Kit]. RNA concentration was determined on a NanoDrop microvolume spectrophotometer (ThermoFisher) and quality was assessed on an Agilent Bioanalyzer 2100.

RNA-seq of blood

The chosen sample sizes and coverage for the bulk RNA-seq were based on empirical data from the prior study (3), which indicated that with 10 animals per group and a two-sided two sample t-test we could detect with a power of ≥ 0.80 and at a significance level of 0.05 a ≥1.5 fold difference in transcription between groups for a given gene. We also were guided by the simulations calculations of Hart et al. (36), which indicated for a biological coefficient of variation of 0.4 within a group, a minimum depth of coverage of ≥ 10, and a target of ≥ 2x fold change that a sample size of 7-8 was sufficient for 80% power and type I error of 5%. For the P. leucopus and M. musculus samples production of cDNA libraries was with the Illumina TruSeq Stranded mRNA kit. After normalization and multiplexing, the libraries were sequenced at the University of California Irvine’s Genomic High Throughput Facility on a Illumina NovaSeq 6000 instrument with paired-end chemistry and 150 cycles to achieve ∼100 million reads per sample for the combined P. leucopus-M. musculus experiment. The same method for producing cDNA libraries was used for the R. norvegicus RNA and the P. leucopus in the infection study, but these were sequenced on a Illumina HiSeq 4000 instrument with paired-end chemistry and 100 cycles. The quality of sequencing reads was analyzed using FastQC (Babraham Bioinformatics). The reads were trimmed of low-quality reads (Phred score of <15) and adapter sequences, and corrected for poor-quality bases using Trimmomatic (12).

For the combined species experiment the mean (95% CI) number of PE150 reads per animal after trimming for quality was 1.1 (1.0-1.2) x 108 for P. leucopus and 1.1 (1.0-1.2) x 108 for M. musculus (p = 0.91). For P. leucopus of this experiment a mean of 83% of the reads mapped to the genome transcript reference set of 54,466; mean coverages for all transcripts and for the mean 62% of reference transcripts with ≥ 1x coverage were 97x and 157x, respectively. For M. musculus of this experiment a mean 91% of the reads mapped to the genome transcript reference set of 130,329; mean coverages for all transcripts and for the mean of 21% of reference transcripts with ≥ 1x coverage were 103x and 568x, respectively. For the lower dose-longer duration experiment with P. leucopus the mean number of PE150 reads was 2.5 (2.3-2.6) x 107. For the rat experiment the mean number of PE100 reads was 2.4 (2.2-2.5) x 107. For the B. hermsii infection experiment the mean number of PE100 reads was 4.9 (4.5-5.3) x 107.

Batched fastq files were subjected to analysis with CLC Genomics Workbench version 23 (Qiagen). Library size normalization was done by the TMM (trimmed mean of M values) method of Robinson and Oshlack (77). The reference genome transcript sets on GenBank were the following: GCF_004664715.2 for P. leucopus LL stock, GCF_000001635.27_GRCm39 for M. musculus C57Bl/6, and GCF_015227675.2_mRatBN7.2 for R. norvegicus. The settings for PE150 reads were as follows for both strands: length fraction of 0.35, similarity fraction of 0.9, and costs for mismatch, insertion, or deletion of 3. For PE100 reads the settings were the same except for length fraction of 0.4. Principal Component Analysis was carried with the “PCA for RNA-Seq” module of the suite of programs.

For the P. leucopus RNA-seq analysis there were 54,466 reference transcripts, of which 48,164 (88%) were mRNAs with protein coding sequences, and 6,302 were identified as non-coding RNAs (ncRNA). Of the 48,164 coding sequences, 40,247 (84%) had matching reads for at least one of the samples. The five most highly represented P. leucopus coding sequences among the matched transcripts of whole blood among treated and control animals were hemoglobin subunits alpha and beta, the calprotectin subunits S100A8 and S100A9, and ferritin heavy chain. For the M. musculus analysis there were available 130,329 reference transcripts: 92,486 (71%) mRNAs with protein coding sequences and 37,843 ncRNAs. Of the coding sequences, 59,239 (64%) were detectably transcribed in one or both groups by the same criterion. The five most highly represented coding sequences of mRNAs of identified genes for M. musculus were hemoglobin subunits alpha and beta, aminolevulinic synthase 2, ferritin light polypeptide 1, and thymosin beta. For R. norvegicus there were 99,126 reference transcripts, of which 74,742 (75%) were mRNAs. The five most highly represented coding sequences of mRNAs of identified genes for R. norvegicus were hemoglobin subunits alpha and beta, beta-2 microglobulin, ferritin heavy chain, and S100a9.

Genome-wide differential gene expression

Differential expression between experimental conditions was assessed with an assumption of a negative binomial distribution for expression level and a separate Generalized Linear Model for each (59). Fold changes in TPM (transcripts per million) were log2-transformed. The False Discovery Rate (FDR) with corrected p value was estimated by the method of Benjamini and Hochberg (10). To assess the limit of detection for differentially expressed genes between 10 animals treated with LPS and 10 with saline alone, we took the data for 4650 reference transcripts for which the mean TPM across 20 P. leucopus was >10 and randomly permuted the data to achieve another 9 sets and calculated the fold-change of sub-groups of 10 and 10 with one random group serving as the proxy of the experimental treatment and other second as the control for each of the sets. The expectation was that mean fold-change of the 9 permuted sets and the 4650 reference sequences would be ∼1. The result was a mean and median of 1.08 with a 99.9% asymmetric confidence interval for the mean of 0.81-1.49. This was an indication that the choices for sample sizes were realistic for achieving detection of ≥ 1.5x fold changes.

Gene Ontology term analysis

M. musculus was selected as the closest reference for the P. leucopus data. The analysis was implemented for data for differentially expressed genes meeting the criteria of a FDR p-value ≤ 0.01 and fold-change of ≥ 1.5. The analysis was implemented with the tools of Metascape (https://metascape.org) (99). Functional enrichment analysis was carried out first with the hypergeometric test and FDR p-value correction (10). Then pairwise similarities between any two enriched terms were computed based on a Kappa-test score (22). Similarity matrices were then hierarchically clustered and a 0.3 similarity threshold was applied to trim resultant trees into separate clusters. The lower the p-value, the less the likelihood the observed enrichment is due to randomness (98). The lowest p-value term represented each cluster shown in the horizontal bar graph. Besides the terms beginning with “GO” and referring to the Gene Ontology resource (http://geneontology.org) (2), others refer to Kegg Pathway database (https://www.kegg.jp) for “mmu..” designations, WikiPathways database (https://www.wikipathways.org) for “WP…” designations, and Reactome database (https://reactome.org) for “R-MMU…” designations.

Targeted RNA-seq

RNA-seq of selected set of protein coding sequences (CDS), which are listed below, was carried out using CLC Genomics Workbench v. 23 (Qiagen). Paired-end reads were mapped with a length fraction of 0.35 for ∼150 nt reads and 0.40 for ∼100 nt reads, a similarity fraction of 0.9, and costs of 3 for mismatch, insertion, or deletion to the CDS of sets of corresponding orthologous mRNAs of P. leucopus, M. musculus, and R. norvegicus. Preliminary expression values were unique reads normalized for total reads across all the samples without adjustment for reference sequence length, as described (3). Exceptions were the endogenous retrovirus coding sequences which differed in lengths between species. For within-and cross-species comparisons we initially normalized three different ways after quality filtering and removing vector and linker sequence: for total reads for the given sample, for unique reads for 12S ribosomal RNA for the mitochondria of nucleated cells in the blood, and for unique reads for the gene Ptprc, which encodes CD45, a marker for both granulocytes and mononuclear cells in the blood. This is described in more detail in Results. Following the recommendation of Hedges et al. we used the natural logarithm (ln) of ratios (37).

The target CDS were as follows: Acod1, Akt1, Akt2, Arg1, Bcl3, Camp, Ccl2, Ccl3, Ccl4, Cd14, Cd177, Cd3d, Cd4, Cd69, Cd8, Cfb, Cgas, Csf1, Csf1r, Csf2, Csf3, Csf3r, Cx3cr1, Cxcl1, Cxcl10, Cxcl2, Cxcl3, Dhx58, Fcer2, Fcgr2a, Fcgr2b, Fcgr3 (CD16), Fgr, Fos, Fpr2, Gapdh, Gbp4, Glrx, Gzmb, Hif1a, Hk3, Hmox1, Ibsp, Icam, Ifih1, Ifit1, Ifng, Il10, Il12, Il18, Il1b, Il1rn, Il2ra, Il4ra, Il6, Il7r, Irf7, Isg15, Itgam, Jak1, Jak2, Jun, Lcn2, Lpo, Lrg, Lrrk2, Ltf, Mapk1, Mmp8, Mmp9, Mpo, MT-Co1, Mt2, Mtor, Mx2, Myc, Myd88, Ncf4, Nfkb1, Ngp, Nos2, Nox1, Nr3c1, Oas1, Olfm4, Padi4, Pbib, Pkm, Ptx, Ptprc (CD45), Retn, Rigi (Ddx58), S100a9, Saa3, Serpine1, Slc11a1 (Nramp), Slpi, Socs1, Socs3, Sod2, Stat1, Stat2, Stat4, Steap1, Sting, Tgfb, Thy1, Timp1, Tlr1, Tlr2, Tlr4, Tnf, Tnfrsf1a, and Tnfrsf9. The sources for these coding sequences were the reference genome transcript sets for P. leucopus, M. musculus, and R. norvegicus listed above. If there were two or more isoforms of the mRNAs and the amino acid sequences differed, the default selection for the coding sequence was the first listed isoform. The lengths of the orthologous pairs of P. leucopus and M. musculus coding sequences were either identical or within 2% of the other. Fcgr1, the gene for high affinity Fc gamma receptor I or CD64, was not included in the comparison, because in P. leucopus it is an untranscribed pseudogene (7). For the targeted RNA-seq of blood of deermice infected with B. hermsii (Table 6), the reference sequence was the cp6.5 plasmid of B. hermsii (NZ_CP015335).

Quantitative PCR assays

Reverse transcriptase (RT)-qPCR assays and the corresponding primers for measurement of transcripts of genes for nitric oxide synthase 2 (Nos2) and glyceraldehyde 3-phosphate dehydrogenase (Gapdh) were those described previously (3). These primers worked for M. musculus as well as P. leucopus using modified cycling conditions. For the arginase 1 transcript assays different primer sets were used for each species. The forward and reverse primer sets for the 352 bp Arg1 product for P. leucopus were 5’-TCCGCTGACAACCAACTCTG and 5’-GACAGGTGTGCCAGTAGATG, respectively. The corresponding primer pairs for a 348 bp Arg1 of M. musculus were 5’-TGTGAAGAACCCACGGTCTG and 5’-ACGTCTCGCAAGCCAATGTA. cDNA synthesis and qPCR were achieved with a Power Sybr Green RNA-to-Ct 1-Step Kit (Applied Biosystems) in 96 MicroAmp Fast Reaction Tubes using an Applied Biosystems StepOne Plus real-time PCR instrument. The initial steps for all assays were 48 °C for 30 min and 95 °C for 10 min. For Arg1 and Nos2 assays, this was followed by 40 cycles of a 2-step PCR of, first, 95°C for 15 s and then, second, annealing and extension at 60 °C for 1 min. The cycling conditions for Gapdh were 40 cycles of 95 °C for 15 s followed by 60 °C for 30 s. Quantitation of genome copies of B. hermsii in extracted DNA was carried out by probe-based qPCR as described (6).

Additional statistics

Means are presented with asymmetrical 95% confidence intervals (CI) to accommodate data that was not normally distributed. Parametric (t test) and non-parametric (Mann-Whitney) tests of significance were 2-tailed. Unless otherwise stated, the t test p value is given. Categorical variables were assessed by 2-tailed Fisher’s exact test. FDR correction of p values for multiple testing was by the Benjamini-Hochberg method (10), as implemented in CLC Genomics Workbench (see above), or False Discovery Rate Online Calculator (https://tools.carbocation.com/FDR). Discriminant Analysis, linear regression, correlation, coefficient of determination, and General Linear Model analyses were performed with SYSTAT v. 13.1 software (Systat Software, Inc.). Box plots with whiskers display the minimum, first quartile, median, third quartile, and maximum.

Data availability

The experiments reported here are associated with National Center for Biotechnology Information (https://ncbi.nlm.nih.gov) BioProjects PRJNA975149 for the combined P. leucopus and M. musculus experiment, PRJNA874306 for the 12 h-lower dose P. leucopus experiment, PRJNA973677 for the R. norvegicus LPS experiment, and PRJNA508222 for the B. hermsii infection of P. leucopus experiment. PRJNA975149 includes 40 BioSamples (SAMN35347136-SAMN35347175) and 40 sets of Sequence Read Archive (SRA) fastq files of Illumina paired-end chemistry reads 1 and 2 (SRR24733648-SRR24733687). PRJNA874306 includes 6 BioSamples (SAMN30561752-SAMN30561757) and the corresponding SRA files SRR24451178-SRR24451180, SRR24451183, SRR24451194, and SRR24451195. PRJNA973677 includes 16 BioSamples (SAMN35351370-SAMN35351385) and the corresponding SRA files SRR24731678-SRR24731693. PRJNA508222 includes 7 BioSamples (SAMN10522571-SAMN10522573; SAMN10522575-SAMN10522578) and SRA files (SRR8283809; SRR8283811-SRR8283816). The datasets of the genome-wide RNA-seq and the differentially-expressed gene analysis of RNA-seq data for the 4 h LPS experiments for P. leucopus, M. musculus, and R. norvegicus are available under the title “Differentially-expressed genes in blood in response to lipopolysaccharide in three rodent species” at the public data repository Dryad (https://datadryad.org). The link for Dryad Tables D1-D6 and their descriptions is https://doi.org/10.5061/dryad.zw3r228dh. A Transcriptome Shotgun Assembly (TSA) for stranded mRNA of blood of the 20 female and male P. leucopus of the 4 h LPS experiment comprised 14,982 contigs of ≥ 400 nt and ≥ 20x coverage and have TSA accession number GKOE00000000.

Supporting information